---
title: "Pacchetto"
author: "Giacomo Colasurdo"
date: "2026-07-21"
output: pdf_document
---

# Sample depth

# Evaluates the temporal sample replication for a single ring-width chronology 
# (RWI/RWL matrix). It automatically detects tree IDs using dplR::autoread.ids() 
# to count unique trees per year (applying a robustness threshold of >= 6 trees). 
# If tree identification fails, it defaults to counting raw series with a 
# threshold of >= 12 series.

```{r}
get_sample_depth <- function(RWI) {
  # 1. Prova a leggere gli ID degli alberi
  ids <- tryCatch(autoread.ids(RWI), error = function(e) NULL)
  
  # 2. Prepariamo il conteggio anno per anno
  coverage_df <- data.frame(
    Anno = as.numeric(rownames(RWI)),
    n_presenti = rowSums(!is.na(RWI))
  )
  
  # 3. Logica della soglia
  if (!is.null(ids)) {
    # Se abbiamo gli ID, la soglia è 6 alberi. 
    # Dobbiamo però capire quante SERIE servono per avere 6 alberi.
    # In media si assume 2 core per albero, quindi 12 serie, 
    # ma usiamo il parametro "6 alberi" come riferimento teorico.
    soglia_val <- 6 
    # Nota: per semplicità di grafico, se contiamo gli alberi attivi per anno:
    n_alberi_per_anno <- sapply(1:nrow(RWI), function(i) {
      length(unique(ids$tree[which(!is.na(RWI[i, ]))]))
    })
    coverage_df$n_valore <- n_alberi_per_anno
    label_y <- "Number of trees"
  } else {
    # Fallback: se autoread.ids fallisce, usiamo 12 serie
    soglia_val <- 12
    coverage_df$n_valore <- coverage_df$n_presenti
    label_y <- "Number of series"}
  
  return(list(df = coverage_df, soglia = soglia_val, label = label_y))}
```

# Reads an individual site's Tucson-formatted file (.rwl) and determines its 
# annual validity based on two criteria:
# - Supera_Soglia: Reaches the minimum robustness threshold (6 trees or 12 series).
# - In_Common_Interval: Falls within the site's 100% common overlap period 
#   calculated via dplR::common.interval().

```{r}
analizza_sito <- function(stand_id) {
  # Costruisci il percorso del file (modifica se necessario)
  file_path <- paste0("File/", stand_id)
  
  # 1. Caricamento sicuro
  rwl <- tryCatch(read.rwl(file_path, format="tucson"), error = function(e) return(NULL))
  if(is.null(rwl)) return(NULL)
  
  anni <- as.numeric(rownames(rwl))
  
  # 2. Calcolo Soglia (6 alberi o 12 serie)
  ids <- tryCatch(autoread.ids(rwl), error = function(e) NULL)
  n_target <- if(!is.null(ids)) 6 else 12
  
  # Per ogni anno, conta quante serie/alberi ci sono
  presenza_anno <- data.frame(Anno = anni)
  if(!is.null(ids)) {
    presenza_anno$count <- sapply(1:nrow(rwl), function(i) {
      length(unique(ids$tree[which(!is.na(rwl[i, ]))]))
    })
  } else {
    presenza_anno$count <- rowSums(!is.na(rwl))}
  
  # 3. Identificazione Common Interval (100% delle serie presenti)
  # common.interval restituisce il sotto-periodo dove TUTTE le serie del file sono presenti
  ci_rwl <- tryCatch(common.interval(rwl, make.plot = FALSE), error = function(e) NULL)
  anni_ci <- if(!is.null(ci_rwl)) as.numeric(rownames(ci_rwl)) else NULL
  
  # Risultato: per ogni anno del sito, segnamo se supera le due soglie
  res <- presenza_anno %>%
    mutate(
      Supera_Soglia = count >= n_target,
      In_Common_Interval = Anno %in% anni_ci)
  
  return(res)}
```

# Iterates over a vector of site IDs, runs 'analizza_sito()' for each, and 
# computes the cumulative relative frequency (%) of available sites across a 
# defined time window (e.g., 1900-2024). Returns a tidy long-format dataframe 
# comparing the availability rate for both inclusion criteria over time.

```{r}
prepara_df_disponibilita <- function(stand_ids, start_year = 1900, end_year = 2024) {
  
  # 1. Storage initialization
  anni_globali <- start_year:end_year
  storage_soglia <- setNames(numeric(length(anni_globali)), anni_globali)
  storage_common <- setNames(numeric(length(anni_globali)), anni_globali)
  siti_letti_con_successo <- 0
  
  # 2. Loop over sites
  for (sid in stand_ids) {
    # Call to your analizza_sito function (must be defined in the global environment)
    dati_sito <- analizza_sito(sid) 
    
    if(!is.null(dati_sito)) {
      siti_letti_con_successo <- siti_letti_con_successo + 1
      
      # Extract valid years within range
      anni_ok_soglia <- as.character(dati_sito$Anno[dati_sito$Supera_Soglia & 
                                                    dati_sito$Anno >= start_year & 
                                                    dati_sito$Anno <= end_year])
      
      anni_ok_common <- as.character(dati_sito$Anno[dati_sito$In_Common_Interval & 
                                                    dati_sito$Anno >= start_year & 
                                                    dati_sito$Anno <= end_year])
      
      # Update accumulator
      storage_soglia[anni_ok_soglia] <- storage_soglia[anni_ok_soglia] + 1
      storage_common[anni_ok_common] <- storage_common[anni_ok_common] + 1}}
  
  # 3. Relative frequency calculation and pivot
  df_long <- data.frame(
    Anno = anni_globali,
    Soglia_6_Alberi = (storage_soglia / length(stand_ids)) * 100,
    Common_Interval_100 = (storage_common / length(stand_ids)) * 100
  ) %>% 
    filter(Soglia_6_Alberi > 0 | Common_Interval_100 > 0) %>%
    pivot_longer(cols = -Anno, names_to = "threshold_Type", values_to = "Relative_Frequency")
  
 # Print a short report to console
  cat("Elaboration completed.\nTotal Site:", length(stand_ids), 
      "\nSite read successfully:", siti_letti_con_successo, "\n")
  
  return(df_long)}
```

```{r}
df_long_finale <- prepara_df_disponibilita(stand_ids = veg$stand_id)

# Ora controlla se il dataframe ha dei dati
head(df_long_finale)

ggplot(df_long_finale, aes(x = Anno, y = Frequenza_Relativa, color = Tipo_Soglia)) +
  # Solo le linee, senza geom_area
  geom_line(linewidth = 1.3) + 
  
  # Scala Y 0-100%
  scale_y_continuous(
    limits = c(0, 100), 
    breaks = seq(0, 100, 10),
    labels = function(x) paste0(x, "%")
  ) +
  
  # Colori puliti
  scale_color_manual(
    values = c("Soglia_6_Alberi" = "lightsalmon", "Common_Interval_100" = "steelblue"),
    labels = c("100% Series (Common Interval)", "Min. 6 Trees (Robust Threshold)")
  ) +
  
  labs(
    title = "Relative Frequency of Site Availability",
    subtitle = paste("Based on", length(unique(veg$stand_id)), "valid sites"),
    x = "Year", 
    y = "Percentage of Total Sites",
    color = "Inclusion Criteria"
  ) +
  
  # Tema bianco senza griglia minore per massima pulizia
  theme_bw() + 
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    panel.grid.minor = element_blank(),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )
ggsave("sample_death_cumulative.pdf", width = 9, height = 6, dpi = 300)
```

# download rwl series

# Connects to the NOAA ITRDB server for European tree-ring measurements, 
# parses the HTML directory listing, filters for valid '.rwl' dataset files, 
# and optionally downloads them to a specified local directory.

```{r}
download_files = function(url,download=TRUE) {
  # Fetch the directory listing HTML
  response = GET(url)
  # Check if the request was successful
  if (status_code(response) != 200) {
    stop("Failed to retrieve the directory listing.")}
  # Parse the HTML content
  page = content(response, "text")
  parsed_html = read_html(page)
  # Extract all anchor (<a>) elements, this includes the rwl and the txt
  links = parsed_html %>% html_nodes("a") %>% html_attr("href")
  # Filter out parent directory links and non-rwl file links
  file_links = links[grepl("\\.rwl", links)]
  # Delete the element ".rwl" from the string
  file_links_spill = str_split(file_links, ".rwl") #Split the string
  int=c()
  for (i in 1:length(file_links_spill)){
    # delete the tail and pick up only the name 
    int=c(int,file_links_spill[[i]][1])}
    final.list=c()
    for (i in int){
      # If the string end with number, have to record it
      if (grepl("\\s*[[:digit:]]$",i)==TRUE){
        final.list=c(final.list,paste0(i, ".rwl"))}}
  int=final.list
  # Download the file list
  if (download){
    for (file_link in int) {
      # Construct the full URL for each file
      file_url = paste0(url, file_link)
      # Define the local destination file path
      destfile = file.path("File", basename(file_link))
      # Download the file
      download.file(file_url, destfile, mode = "wb")
      # Print status
      cat("Downloaded:", file_link, "\n")}}
  return(final.list)}
```

# Stand Processing

# Wrapper function that extracts stand metadata (coordinates, 
# elevation, species) by cross-referencing official ITRDB metadata 
# and raw file header information parsed via ecfh().

```{r}
ec <- function(file_path,metadata_path){
  result <- list(metadata = list(), header = list())
  metadata <- read.delim(metadata_path)
  nome_stand_up <- toupper(gsub("\\.rwl$", "", basename(file_path), ignore.case = TRUE))
  if (nome_stand_up %in% metadata$ITRDB_Code) {
    coordinate <- metadata[metadata$ITRDB_Code==nome_stand_up,c("Latitude","Longitude")]
    info_stand <- metadata[metadata$ITRDB_Code==nome_stand_up,c("Elevation","Species")]
    coor <- c(coordinate$Longitude, coordinate$Latitude)
    result$metadata$coordinate <- coor
    result$metadata$info <- info_stand}
  coordinate_estratte <- ecfh(file_path,lista_specie=unique(metadata$Species))
  if (!is.null(coordinate_estratte$coord_Y) && 
      !is.null(coordinate_estratte$coord_X)){
    coor_extr <- c(coordinate_estratte$coord_Y,coordinate_estratte$coord_X)
    result$header$coordinate <- coor_extr}
  if (!is.null(coordinate_estratte$Specie)){
    result$header$specie <- coordinate_estratte$Specie}
  result$header$head <- coordinate_estratte$header
  return(result)}
```

# Parses the first 3 lines (header) of a raw Tucson (.rwl) file. 
# Extracts 4-letter species codes and uses multiple regex pattern 
# matches to parse and standardize geographic coordinates (lon/lat)

```{r}
ecfh <- function(file_path,lista_specie) {
  result <- list()
  header_lines <- tryCatch({
    readLines(file_path, n = 3, warn = FALSE)
  }, error = function(e) {
    warning("Cannot read file: ", file_path)
    return(NULL)})
  if (is.null(header_lines)) return(result)
  header_text <- str_squish(paste(header_lines, collapse = " "))
  result$header <- header_text
  # Estrazione Specie
  tutte_quattro <- unlist(str_extract_all(header_text, "\\b[A-Z]{4}\\b"))
  if (length(tutte_quattro) == 1) result$Specie <- tutte_quattro
  match_lista <- intersect(tutte_quattro, lista_specie)
  if (length(match_lista) == 1) result$Specie <- match_lista
  # --- PATTERN 1 (coordinates in "M N E" format) ---
  m1 <- str_match(header_text, "(-?[[:digit:]]+)\\s*M\\s*(-?[[:digit:]]+)\\s*N?(-?[[:digit:]]+)\\s*E?")
  if (!is.na(m1[1, 1])) {
    result$coord_X <- as.numeric(m1[1, 3]) / 100
    result$coord_Y <- as.numeric(m1[1, 4]) / 100
    return(result)}
  # --- PATTERN 4 (signed DDMM-DDDMM format) ---
  # Cerca un segno (+ o -) seguito da cifre, poi un altro segno e altre cifre
  m4 <- str_match(header_text, "([+-][[:digit:]]{3,5})([+-][[:digit:]]{3,6})")
  if (!is.na(m4[1, 1])) {
    result$coord_X <- as.numeric(m4[1, 2]) / 100
    result$coord_Y <- as.numeric(m4[1, 3]) / 100
    return(result)}
  # --- PATTERN 2 (decimal coordinates followed by underscores) ---
  m2 <- str_match(header_text, "(-?[[:digit:]]+\\.?[[:digit:]]*)\\s+(-?[[:digit:]]+\\.?[[:digit:]]*)\\s+_{1,}")
  if (!is.na(m2[1, 1])) {
    val2 <- as.numeric(m2[1, 2])
    val3 <- as.numeric(m2[1, 3])
    if (val2 < 1000) {
      result$coord_X <- val3
      result$coord_Y <- val2
    } else {
      result$coord_X <- val2 / 100
      result$coord_Y <- val3 / 100}
    return(result)}
  # --- PATTERN 3 ("N ... W" format) ---
  m3 <- str_match(header_text, "\\s*(-?[[:digit:]]+)\\s*N(-?[[:digit:]]+)\\s*W")
  if (!is.na(m3[1, 1])) {
    result$coord_X <- as.numeric(m3[1, 1]) / 100
    result$coord_Y <- as.numeric(m3[1, 2]) / 100
    return(result)}
  # --- PATTERN 10 (fixed-width 10-character format) ---
  m10 <- str_match(header_text, "(.{10})\\s+_")
  if (!is.na(m10[1, 1])) {
    raw_val <- m10[1, 2]
    result$coord_X <- as.numeric(substr(raw_val, 1, 5)) / 100
    result$coord_Y <- as.numeric(substr(raw_val, 6, 10)) / 100
    return(result)}
  return(result)}
```

# Evaluates ring-width series to identify a continuous common 
# time interval satisfying minimum tree count and period length constraints. 
# Falls back to column-wise counts if tree grouping fails.

```{r}
common_interval <- function(rwl, tim, min.trees = 6, start.yr = 1901, 
                                    end.yr = 2025, min.period.len = 9) {
  if (!is.data.frame(rwl)) stop("'rwl' deve essere un data.frame")
  yrs <- as.numeric(row.names(rwl))
  rwlNotNA <- !is.na(rwl)
  samp.depth <- tryCatch({
    tim <- autoread.ids(rwl)
    apply(rwlNotNA, 1, function(row_exists) {
      series_present <- names(rwl)[row_exists]
      unique_trees <- unique(tim$tree[rownames(tim) %in% series_present])
      length(unique_trees)})
  }, error = function(e) {
    message("Nota: Impossibile raggruppare per albero (nomi serie non standard). Uso il conteggio colonne.")
    min.trees <- min.trees*2
    return(rowSums(rwlNotNA))})
  potential.years <- samp.depth >= min.trees & yrs >= start.yr & yrs <= end.yr
  runs <- rle(potential.years)
  max.run.len <- max(runs$lengths[runs$values == TRUE], 0)
  output <- list(found = FALSE, rwl = NULL)
  if (max.run.len >= min.period.len) {
    end.indices <- cumsum(runs$lengths)
    start.indices <- c(1, end.indices[-length(end.indices)] + 1)
    valid.runs <- which(runs$values == TRUE & runs$lengths >= min.period.len)
    target.run <- valid.runs[which.max(runs$lengths[valid.runs])]
    row.indices <- start.indices[target.run]:end.indices[target.run]
    res <- rwl[row.indices, , drop = FALSE]
    keep.col.mask <- colSums(!is.na(res)) > 0
    output$found <- TRUE
    output$rwl <- res[, keep.col.mask, drop = FALSE]}
  return(output)}
```

# Iterates through all .rwl files in a directory, applies spline 
# detrending, isolates common time windows, and builds standard mean chronologies.

```{r}
elabora_cronologie_rwl <- function(dir_file) {
  
  nomi_file <- list.files(dir_file, pattern = "\\.rwl$", full.names = FALSE)
  totale_stands <- length(nomi_file)
  results_base <- list()
  
  for (i in seq_along(nomi_file)) {
    stand <- nomi_file[i]
    file_path <- file.path(dir_file, stand)
    
    cron_df <- tryCatch({
      RWL  <- read.rwl(file_path, format = "tucson")
      DET  <- detrend(RWL, method = "Spline", nyrs = 30, f = 0.5)
      det  <- common_interval(DET)
      if (!det$found) stop("Intervallo comune non trovato")
      cron <- chron(det$rwl, biweight = TRUE, prewhiten = TRUE)
      data.frame(
        anno = as.numeric(rownames(cron)), 
        rwl  = as.numeric(cron[, 1]))},
      error = function(e) NULL)
    
    if (!is.null(cron_df)) {
      results_base[[stand]] <- cron_df}
    
    if (i %% 100 == 0) cat("Processati:", i, "/", totale_stands, "\n")}
  return(results_base)}
```

# Metadata 

# Consolidates stand-level metadata (coordinates, altitude, species) 
# by combining header extraction and official ITRDB text records. 
# Outputs a consolidated master dataframe and a SpatVector object for spatial analysis.

```{r}
estrai_metadati_e_master <- function(stand_validi, path_metadata, dir_file) {
  metadata_txt <- read.delim(path_metadata, stringsAsFactors = FALSE)
  lista_master <- list()
  
  for (i in seq_along(stand_validi)) {
    stand <- stand_validi[i]
    st_id_clean <- gsub("\\.rwl$", "", stand, ignore.case = TRUE)
    file_path <- file.path(dir_file, stand)
    
    ec_res <- tryCatch({
      ec(file_path, path_metadata)}, 
      error = function(e) NULL)
    
    if (is.null(ec_res)) next
    
    # --- COORDINATE GREZZE ---
    h_coords  <- ec_res$header$coordinate
    lon_ext   <- if (!is.null(h_coords)) h_coords[1] else NA
    lat_ext   <- if (!is.null(h_coords)) h_coords[2] else NA
    
    m_coords  <- ec_res$metadata$coordinate
    lon_itrdb <- if (!is.null(m_coords)) m_coords[1] else NA
    lat_itrdb <- if (!is.null(m_coords)) m_coords[2] else NA
    
    lon_def <- ifelse(!is.na(lon_ext), lon_ext, lon_itrdb)
    lat_def <- ifelse(!is.na(lat_ext), lat_ext, lat_itrdb)
    
    # --- ALTITUDINE ---
    idx_meta <- which(toupper(metadata_txt$ITRDB_Code) == toupper(st_id_clean))
    elev <- if (length(idx_meta) > 0) metadata_txt$Elevation[idx_meta[1]] else NA
    
    # --- SPECIE ---
    sp_header <- if (!is.null(ec_res$header$specie)) toupper(trimws(ec_res$header$specie)) else NA
    sp_meta   <- if (length(idx_meta) > 0) trimws(metadata_txt$Species[idx_meta[1]]) else NA
    sp_finale <- ifelse(!is.na(sp_meta), sp_meta, ifelse(!is.na(sp_header), sp_header, NA))
    
    # --- HEADER TEXT ---
    head_txt <- if (!is.null(ec_res$header$head)) paste(ec_res$header$head, collapse = " ") else NA
    
    # --- COMPILAZIONE RECORD UNICO ---
    lista_master[[stand]] <- data.frame(
      stand_id        = stand,
      lon_extract     = as.numeric(lon_ext),
      lat_extract     = as.numeric(lat_ext),
      lon_ITRDB       = as.numeric(lon_itrdb),
      lat_ITRDB       = as.numeric(lat_itrdb),
      lon_definitiva  = as.numeric(lon_def),
      lat_definitiva  = as.numeric(lat_def),
      altitude        = as.numeric(elev),
      species         = as.character(sp_finale),
      header_text     = as.character(head_txt),
      stringsAsFactors = FALSE)
    
    if (i %% 100 == 0) cat("Metadati estratti:", i, "/", length(stand_validi), "\n")}
  
  # Assemblaggio Dataframe Unico
  df_master_totale <- do.call(rbind, lista_master)
  
  # Creazione SpatVector basato sulle coordinate definitive
  df_valid_geo <- df_master_totale[!is.na(df_master_totale$lon_definitiva) & 
                                   !is.na(df_master_totale$lat_definitiva), ]
  punti_vect <- vect(
    df_valid_geo, 
    geom = c("lon_definitiva", "lat_definitiva"), 
    crs = "+proj=longlat +datum=WGS84")
  
  return(list(
    df_master  = df_master_totale,
    punti_vect = punti_vect))}
```

# Coordinate solution

# Identifies tree-ring stands whose coordinates incorrectly fall 
# into marine cells (NA values in the CRU raster). For these problematic 
# stands, it performs a localized search (+/- 2.5 degrees) to snap the point 
# to the nearest valid terrestrial grid cell.

```{r}
gestisci_punti_in_acqua <- function(df_audit, path_raster_cru = "dati/CRU_TS_pre.nc") {
  
  # 1. Carichiamo il raster CRU
  raster_cru <- rast(path_raster_cru)[[1]]
  
  # 2. Definiamo inizialmente lon/lat_definitiva usando la miglior coordinata disponibile
  # (Se lon_extract è valida usiamo quella, altrimenti ITRDB)
  df_audit$lon_definitiva <- ifelse(!is.na(df_audit$lon_extract), df_audit$lon_extract, df_audit$lon_ITRDB)
  df_audit$lat_definitiva <- ifelse(!is.na(df_audit$lat_extract), df_audit$lat_extract, df_audit$lat_ITRDB)
  
  # 3. Creiamo lo SpatVector basato ESCLUSIVAMENTE sulle coordinate DEFINITIVE
  punti_def_vect <- vect(
    df_audit[!is.na(df_audit$lon_definitiva) & !is.na(df_audit$lat_definitiva), ], 
    geom = c("lon_definitiva", "lat_definitiva"), 
    crs = "+proj=longlat +datum=WGS84")
  
  # 4. Estraiamo i valori dal raster CRU per le coordinate DEFINITIVE
  valori_cru <- terra::extract(raster_cru, punti_def_vect, ID = FALSE)[, 1]
  
  # 5. Isoliamo gli stand la cui coordinata DEFINITIVA cade su NA (in acqua)
  idx_acqua <- which(is.na(valori_cru))
  stand_in_acqua_ids <- punti_def_vect$stand_id[idx_acqua]
  
  # Selezioniamo solo le righe degli stand finiti in mare
  df_res <- df_audit[df_audit$stand_id %in% stand_in_acqua_ids, ]
  
  # 6. SNAP-TO-LAND: Spostiamo i punti in acqua sulla cella di terraferma CRU più vicina
  punti_acqua_vect <- punti_def_vect[idx_acqua]
  
# Inizializziamo i vettori per le nuove coordinate snapped
  lons_snap <- numeric(length(idx_acqua))
  lats_snap <- numeric(length(idx_acqua))
  
  # 5. SNAP-TO-LAND LOCALE (Ciclo punto per punto su raggio ristretto)
  for (j in 1:length(idx_acqua)) {
    p_singolo <- punti_acqua_vect[j]
    
    # Definiamo una finestra locale di ricerca (+/- 2.5 gradi attorno al punto)
    ext_locale <- ext(crds(p_singolo)[1] - 2.5, crds(p_singolo)[1] + 2.5,
                      crds(p_singolo)[2] - 2.5, crds(p_singolo)[2] + 2.5)
    
    # Ritagliamo il raster sulla finestra locale ed estraiamo le sole celle di TERRA
    r_locale <- crop(raster_cru, ext_locale)
    celle_terra_locali <- as.points(r_locale, values = FALSE, na.rm = TRUE)
    
    # Se troviamo celle terrestri nella finestra locale, prendiamo la più vicina
    if (length(celle_terra_locali) > 0) {
      idx_nearest <- nearest(p_singolo, celle_terra_locali)$to_id
      coords_nearest <- crds(celle_terra_locali[idx_nearest])
      
      lons_snap[j] <- coords_nearest[1]
      lats_snap[j] <- coords_nearest[2]
    } else {
      # Fallback di sicurezza nel caso remoto che non trovi terra entro 2.5 gradi
      lons_snap[j] <- crds(p_singolo)[1]
      lats_snap[j] <- crds(p_singolo)[2]}}
  
  # Assegniamo le coordinate snapped corrette
  df_res$lon_snap <- lons_snap
  df_res$lat_snap <- lats_snap
  
  # Selezioniamo le colonne
  audit_completo <- df_res[, c("stand_id",
                               "lon_extract", "lat_extract", 
                               "lon_ITRDB", "lat_ITRDB", 
                               "lon_definitiva", "lat_definitiva",
                               "lon_snap", "lat_snap",
                               "header_text")]
  
  return(audit_completo)}
```

# Merges corrected marine stands back into the full dataset. 
# It applies either a longitude sign inversion (-Lon, common header typo) 
# or the local Snap-To-Land coordinate based on a logical decision vector.

```{r}
applica_correzioni_coordinate <- function(df_audit_totale, df_audit_acqua, vec_inversione) {
  # 1. Copia del dataframe completo di tutti gli stand
  df_corr <- df_audit_totale
  # Definiamo le coordinate iniziali (Header con fallback su ITRDB)
  df_corr$lon_corretta <- ifelse(!is.na(df_corr$lon_extract), df_corr$lon_extract, df_corr$lon_ITRDB)
  df_corr$lat_corretta <- ifelse(!is.na(df_corr$lat_extract), df_corr$lat_extract, df_corr$lat_ITRDB)
  df_corr$metodo_correzione <- "Nessuna_Correzione"
  
  # 2. Identifichiamo gli stand finiti in mare
  stand_in_acqua <- df_audit_acqua$stand_id
  # Mappatura del vettore di inversione (-Lon) per gli stand in acqua
  inverte_segno <- vec_inversione[stand_in_acqua]
  inverte_segno[is.na(inverte_segno)] <- FALSE
  
  # 3. Calcolo delle nuove coordinate solo per gli stand in mare
  lon_rettificate <- ifelse(inverte_segno, -1 * df_audit_acqua$lon_definitiva, df_audit_acqua$lon_snap)
  lat_rettificate <- ifelse(inverte_segno, df_audit_acqua$lat_definitiva, df_audit_acqua$lat_snap)
  metodi_applicati <- ifelse(inverte_segno, "Inversione_Segno", "Snap_To_Land")
  
  # 4. Aggiornamento nel dataframe completo
  idx <- match(stand_in_acqua, df_corr$stand_id)
  df_corr$lon_corretta[idx] <- lon_rettificate
  df_corr$lat_corretta[idx] <- lat_rettificate
  df_corr$metodo_correzione[idx] <- metodi_applicati
  # 5. Filtro per eliminare eventuali stand completamente privi di coordinate
  df_valid <- df_corr[!is.na(df_corr$lon_corretta) & !is.na(df_corr$lat_corretta), ]
  
  # 6. Creazione dello SpatVector finale COMPLETO  
  names(df_valid)[names(df_valid) == "temp_lon"] <- "longitude"
  names(df_valid)[names(df_valid) == "temp_lat"] <- "latitude"
  punti_geografici_corretti <- vect(
    df_valid,
    geom = c("longitude", "latitude"),
    crs = "+proj=longlat +datum=WGS84")
  return(list(
    vect  = punti_geografici_corretti,
    audit = df_corr))}
```

# Taxonomies

# Parses a raw text file containing taxonomic hierarchy strings 
# and constructs a structured data frame aligned with a target 
# species vector.

```{r}
leggi_tasso <- function(file_txt, vet_specie){
  testo_grezzo <- readChar(file_txt, file.info(file_txt)$size)
  tassi_caricati <- unlist(strsplit(testo_grezzo, "\",\\s*\"|\",\""))
  tassi_caricati <- gsub("^\"|\"$", "", tassi_caricati)
  tassi_caricati <- gsub("[\r\n\t]+", " ", tassi_caricati)
  extract_taxa <- function(string, label) {
    pattern <- paste0(label, "\\s*([A-Za-z\\s\\-\\.]+?)\\s*\\(")
    match <- str_match(string, pattern)
    return(match[, 2]) }
  Tax_Final <- data.frame(
    ID_Specie = unique(vet_specie),
    Kingdom   = sapply(tassi_caricati, extract_taxa, "Kingdom"),
    Phylum    = sapply(tassi_caricati, extract_taxa, "Phylum"),
    Class     = sapply(tassi_caricati, extract_taxa, "Class"),
    Category  = sapply(tassi_caricati, extract_taxa, "Category"),
    Order     = sapply(tassi_caricati, extract_taxa, "Order"),
    Family    = sapply(tassi_caricati, extract_taxa, "Family"),
    Genus     = sapply(tassi_caricati, extract_taxa, "Genus"),
    Species = sapply(tassi_caricati, extract_taxa, "Species"))
  Tax_Final <- as.data.frame(lapply(Tax_Final, trimws))
  return(Tax_Final)}
```

# Object clima creation

# Extracts spatial-temporal data for a single geographical point 
# from a NetCDF raster file using the 'terra' package.

```{r}
extraction <- function(coor, file_nc, nome){
  nc_pet <- rast(file_nc, subds = nome)
  lon <- coor[1]
  lat <- coor[2]
  punto <- vect(matrix(c(lon, lat), ncol=2),)
  pet_ext <- terra::extract(nc_pet, punto, ID = FALSE)
  date_ext <- time(nc_pet)
  data_x = format(date_ext, "%b-%Y")
  df_clima <- data.frame( time = date_ext, data = data_x, var = as.numeric(pet_ext))
  colnames(df_clima)[3] <- nome
  df_finale <- df_clima[,c(2,3)]
  return(df_finale)}
```

# Iterates over unique spatial points (pixels) and extracts multiple 
# climate variables (SPEI, temperature, precipitation, PET) from 
# NetCDF files located in the 'dati/' directory.

```{r}
estrazione_clima <- function(df_unici) {
  file_clima <- c("spei01.nc","CRU_TS_tmp.nc","CRU_TS_tmn.nc",
                  "CRU_TS_pre.nc","CRU_TS_pet.nc")
  nomi <- c("spei","tmp","tmn","pre","pet")
  lista_meteo_raw <- list()
  for(i in 1:nrow(df_unici)) {
    p_id <- df_unici$pixel_id[i]
    coor <- c(df_unici$longitude[i],df_unici$latitude[i])
    df_pixel_accumulato <- NULL
    for(j in 1:length(file_clima)) {
      per <- file.path("dati",file_clima[j])
      df_temp <- extraction(coor, per, nome = nomi[j])
      if (is.null(df_pixel_accumulato)) {df_pixel_accumulato <- df_temp
      } else {
        df_pixel_accumulato <- merge(df_pixel_accumulato, df_temp, 
                                     by = "data", all = TRUE)}}
    lista_meteo_raw[[p_id]] <- df_pixel_accumulato
    if(i %% 50 == 0) cat("Processati", i, "pixel su", nrow(df_unici), "\n")}
  df_pixel_accumulato <- df_pixel_accumulato[order(df_pixel_accumulato$time), ]
  return(lista_meteo_raw)}
```

# metodo più veloce:

```{r}
library(terra)
library(dplyr)

# ==============================================================================
# FUNZIONE 1: ESTRAZIONE SINGOLO PUNTO (Mantiene l'ordine temporale corretto)
# ==============================================================================
extraction <- function(coor, nc_raster, nome_var) {
  # Creazione dello SpatVector punto WGS84
  punto <- vect(matrix(coor, ncol = 2), crs = "+proj=longlat +datum=WGS84")
  
  # Estrazione valori temporali per il punto
  valori <- as.numeric(terra::extract(nc_raster, punto, ID = FALSE))
  
  # Estrazione delle date reali (Date object)
  date_ext <- time(nc_raster)
  
  # Dataframe con data vera (Date) e label formattata
  df_clima <- data.frame(
    time = date_ext,
    data = format(date_ext, "%b-%Y"),
    val  = valori
  )
  colnames(df_clima)[3] <- nome_var
  
  return(df_clima)
}

# ==============================================================================
# FUNZIONE 2: ITERATORE SUI PIXEL (Caricamento ottimizzato dei raster)
# ==============================================================================
estrazione_clima <- function(df_unici, dir_dati = "DendroCrhAn") {
  
  file_clima <- c("spei01.nc", "CRU_TS_tmp.nc", "CRU_TS_tmn.nc", "CRU_TS_pre.nc", "CRU_TS_pet.nc")
  nomi_var   <- c("spei", "tmp", "tmn", "pre", "pet")
  
  cat("Caricamento in memoria delle griglie raster...\n")
  # Carichiamo i raster UNA SOLA VOLTA all'inizio della funzione
  lista_raster <- list()
  for(k in seq_along(file_clima)) {
    path_nc <- file.path(dir_dati, file_clima[k])
    lista_raster[[nomi_var[k]]] <- rast(path_nc)
  }
  
  lista_meteo_raw <- list()
  tot_pixel <- nrow(df_unici)
  
  cat("Inizio estrazione su", tot_pixel, "pixel unici...\n")
  
  for(i in 1:tot_pixel) {
    p_id <- as.character(df_unici$pixel_id[i])
    coor <- c(df_unici$longitude[i], df_unici$latitude[i])
    
    df_pixel_accumulato <- NULL
    
    for(var_name in nomi_var) {
      r_obj <- lista_raster[[var_name]]
      df_temp <- extraction(coor, r_obj, nome_var = var_name)
      
      if (is.null(df_pixel_accumulato)) {
        df_pixel_accumulato <- df_temp
      } else {
        # Unione mantenendo la colonna 'time' (Date) per preservare l'ordine cronologico
        df_pixel_accumulato <- merge(
          df_pixel_accumulato, 
          df_temp[, c("time", var_name)], 
          by = "time", 
          all = TRUE
        )
      }
    }
    
    # Riordino finale garantito per data
    df_pixel_accumulato <- df_pixel_accumulato[order(df_pixel_accumulato$time), ]
    
    # Salvataggio nella lista con chiave pixel_id
    lista_meteo_raw[[p_id]] <- df_pixel_accumulato
    
    if(i %% 50 == 0 || i == tot_pixel) {
      cat("Processati", i, "pixel su", tot_pixel, "\n")
    }
  }
  
  return(lista_meteo_raw)
}
```

# da usare così

```{r}
library(dplyr)
library(terra)

# 1. Caricamento metadati essenziali con i pixel ID
df_veg <- readRDS("DendroCrhAn/meta_essential.rds")

# 2. Isolamento dei pixel unici
df_pixel <- df_veg %>%
  distinct(pixel_id, .keep_all = TRUE) %>%
  select(pixel_id, longitude, latitude)

# 3. Estrazione climatica multivariata
clima <- estrazione_clima(df_pixel, dir_dati = "DendroCrhAn")

# 4. Salvataggio della lista climatica
saveRDS(clima, "DendroCrhAn/lista_clima.Rds")

cat("Salvataggio completato! 'lista_clima.Rds' contiene le serie storiche per", length(clima), "pixel.\n")
```

# rwl series plot

```{r}
scostamento <- function(df, mod = FALSE){
  if (nchar(df[1,1]) != 8) {stop("La prima colonna del dataframe deve contenere le date in formato gen-1901")}
  df_due <- df 
  colnames(df_due[1]) <- "data"
  df_due <- df_due %>%
    mutate (mese = as.factor(str_sub(data, start = 1, end = 3))) %>%
    group_by (mese) %>%
    mutate(across(where(is.numeric), ~ .x - mean(.x, na.rm = TRUE),
                  .names = "dif_{.col}")) %>%
    ungroup()
  if (mod) {df_tre <- df_due %>%
    mutate(across(starts_with("dif_"), abs))
  } else {df_tre <- df_due}
  df_qtr <- df_tre %>%
    mutate(anno = as.numeric(str_sub(data, -4))) %>%
    group_by(anno) %>%
    summarise(across(starts_with("dif_"), ~ sum(.x, na.rm = TRUE),
                     .names = "sum_{.col}")) %>%
    ungroup()
  return(df_qtr)}
```


```{r}
plot_preparation <- function(RWI,spei_df) {
  
  ccf_multiple <- function(s1,altre_serie,nomi) {
    correlazioni_max <- c()
    lag_max <- c()
    for (s_corrente in altre_serie) {
      risultato_ccf <- ccf(s1, s_corrente, plot = FALSE, na.action = na.omit)
      valori_acf <- as.numeric(risultato_ccf$acf)
      valori_lag <- as.numeric(risultato_ccf$lag)
      indice_max <- which.max(abs(valori_acf))
      correlazioni_max <- c(correlazioni_max, round(valori_acf[indice_max],3))
      lag_max <- c(lag_max, valori_lag[indice_max])}
    matrice_risultati <- matrix(c(correlazioni_max, lag_max), nrow = 2, byrow = TRUE)
    df_finale <- as.data.frame(matrice_risultati)
    rownames(df_finale) <- c("Corr", "Lag")
    colnames(df_finale) <- nomi
    return(df_finale)}
  
  rwl <- common_interval(RWI)$rwl
  rwl_bsgc <- common.interval(RWI, make.plot = FALSE)
  det <- detrend(rwl, method = "Spline", nyrs = 30, f = 0.5)
  it <- interval.trend(rwl, trend.thresh = 0, IT.thresh = 95, make.plot = FALSE)
  BSGC <- bsgc(rwl_bsgc, make.plot = FALSE, maxlag = 5)
  n_rep <- length(names(BSGC$neg))
  df_bsgc <- data.frame(year=names(BSGC$neg),Methods=rep("BSGC",n_rep))
  pointer_years_df <- it$out[it$out$nature<0,1,drop=FALSE]
  pointer_years_df$Methods <- rep("IT",nrow(pointer_years_df))
  df_py <- rbind(df_bsgc,pointer_years_df)
  df_py$year <- as.numeric(df_py$year)
  cron <- chron(det, biweight = TRUE, prewhiten = TRUE)
  det <- tibble::rownames_to_column(det, var = "Anno")
  det_long <- det %>%
    tidyr::pivot_longer(
      cols = -c(Anno),
      names_to = "serie",
      values_to = "valore" ) %>% as.data.frame()
  det_long$Anno <- as.numeric(det_long$Anno)
  det_long <- arrange(det_long, serie)
  
  cron_df <- data.frame(
    Anno = as.numeric(rownames(cron)),
    TRW = as.numeric(cron[, 1]))
  min_old <- min(spei_df[,2], na.rm = TRUE)
  max_old <- max(spei_df[,2], na.rm = TRUE)
  spei_df[,2] <- ((spei_df[,2] - min_old) / (max_old - min_old)) + 1.5
  colnames(spei_df) <- c("Anno","SPEI")
  spei_df <- spei_df[spei_df$Anno %in% time(rwl),]
  max_y <- max(c(spei_df$SPEI, cron_df$TRW), na.rm = TRUE)*1.05
  min_y <- min(c(spei_df$SPEI, cron_df$TRW), na.rm = TRUE)*0.8
  df_py <- df_py %>% mutate(
    y_plot = case_when( Methods == "BSGC" ~ 2.60, Methods == "IT"   ~ 2.75))
  corr <- left_join(cron_df,spei_df, by="Anno")
  
  if (nrow(df_py)!=0){
    for (py_meth in unique(df_py$Methods)){
      num <- ncol(corr)+1
      corr[,num] <- rep(0,nrow(corr))
      corr[corr$Anno %in% df_py[df_py$Methods==py_meth,"year"],num] <- 1
      colnames(corr)[num] <- py_meth}
    s1 <- corr[,2]
    if (ncol(corr)<5){altre_serie <- list(s2=corr[,3], s3=corr[,4])
    } else {altre_serie <- list(s2=corr[,3], s3=corr[,4], s4=corr[,5])}
    nomi <- colnames(corr)[c(3:ncol(corr))]
  } else {altre_serie <- list(s2=corr[,3])
  nomi <- colnames(corr)[c(3)]}
  df_corr <- ccf_multiple(s1,altre_serie,nomi)
  df_corr$Misure <- c(max(det_long$Anno)-5,min_y)
  return(list(rwi=det_long, master=cron_df, spei=spei_df, df=df_py, corr_df=df_corr))}

RWI <- read.rwl("File/swit321.rwl")
pix <- df_micro[df_micro$stand_id=="swit321.rwl","pixel_id"]
spei_df <- as.data.frame(scostamento(clima[[pix]][,c(1,2)]))
ogg_swit <- plot_preparation(RWI,spei_df)
```

# Cross correlation 

```{r}
correlazione_significativa_lag <- function(df_tax, sel_orders, max_lag = 4) {
  # 1. Filtro per Genere
  df_lavoro <- df_tax
  if (!is.null(sel_orders)) {
    df_lavoro <- df_lavoro %>% filter(gruppo_specie %in% sel_orders)
  }
  # 2. Funzione interna: Estrae solo i dati POSITIVI e SIGNIFICATIVI
  get_sig_cor_data <- function(id_stand, id_pixel) {
    id_s <- as.character(id_stand); id_p <- as.character(id_pixel)
    if (!(id_s %in% names(crono)) | !(id_p %in% names(clima))) return(NULL)
    tryCatch({
      df_a <- crono[[id_s]]; df_b <- clima[[id_p]]
      df_spei_t <- as.data.frame(scostamento(df_b[, c("data", "spei")]))
      df_m <- merge(df_a, df_spei_t, by = "anno")
      if (nrow(df_m) < 10) return(NULL)
      ccf_res <- ccf(df_m$sum_dif_spei, df_m$rwl, lag.max = max_lag, plot = FALSE, na.action = na.pass)
      n_obs <- sum(complete.cases(df_m))
      soglia <- 1.96 / sqrt(n_obs)
      # MODIFICA CRUCIALE: Teniamo solo cor > soglia (esclude negative e non sig)
      return(data.frame(
        stand_id = id_s,
        lag = abs(as.numeric(ccf_res$lag)), 
        cor = as.numeric(ccf_res$acf)
      ) %>% filter(as.numeric(ccf_res$lag) <= 0, cor > soglia))
    }, error = function(e) return(NULL))}
  # 3. Elaborazione Dati
  res_sig <- df_lavoro %>%
    group_by(gruppo_specie) %>%
    group_modify(~ {
      bind_rows(mapply(get_sig_cor_data, .x$stand_id, .x$pixel_id, SIMPLIFY = FALSE))
    }) %>% ungroup()
  # 4. Calcolo Frequenze (per le colonne sienna)
  n_siti_tot <- df_lavoro %>% group_by(gruppo_specie) %>% summarise(n_tot = n_distinct(stand_id), .groups = 'drop')
  df_frequenze <- res_sig %>%
    group_by(gruppo_specie, lag) %>%
    summarise(freq_ass = n(), .groups = 'drop') %>%
    left_join(n_siti_tot, by = "gruppo_specie") %>%
    mutate(freq_rel = (freq_ass / n_tot) * 100)
  # 5. Parametri Grafici
  my_scale <- 50 
  pos_jitter <- position_jitter(width = 0.25, seed = 123)
  pal <- paletteer::paletteer_d("fishualize::Anisotremus_virginicus")[c(1,3)]
  # 6. Costruzione Grafico
  p <- ggplot(res_sig, aes(x = as.numeric(as.character(lag)), y = cor, color = gruppo_specie)) +
    # Barre (senza jitter)
    geom_col(data = df_frequenze, 
             aes(x = as.numeric(as.character(lag)) + 0.4, y = freq_rel / my_scale),
             inherit.aes = FALSE, fill = pal[2], alpha = 0.5, width = 0.15) +
    # Linee e Punti (stesso jitter, stessi dati)
    geom_linerange(aes(ymin = 0, ymax = cor, y = cor), 
                   position = pos_jitter, alpha = 0.4, linewidth = 0.4) +
    geom_point(position = pos_jitter, size = 1.5, alpha = 0.7) +
    scale_y_continuous(
      name = "Correlation Coefficient (r)",
      expand = expansion(mult = c(0, 0.05)),
      sec.axis = sec_axis(trans = ~ . * my_scale, name = "Significant Responses (%)")
    ) +
    # Definiamo i limiti qui: questo NON elimina i dati
    coord_cartesian(ylim = c(0, 0.8)) +
    scale_x_continuous(breaks = seq(0, max_lag, 1), name = "Lag (Years)") +
    paletteer::scale_color_paletteer_d("fishualize::Anisotremus_virginicus") +
    facet_wrap(~gruppo_specie, ncol = 1) + 
    theme_minimal() +
    labs(
      title = "Species-Specific Patterns of Significant Cross-Correlations"
    ) +
    theme(
      legend.position = "none",
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.line.x = element_line(color = "black"),
      axis.text.x = element_text(color = "black"),
      axis.ticks.x = element_line(color = "black"),
      axis.title.x = element_text(color = "black"),
      axis.line.y.left = element_line(color = pal[1]),
      axis.text.y.left = element_text(color = pal[1]), # Numeri etichette
      axis.ticks.y.left = element_line(color = pal[1]),
      axis.title.y.left = element_text(color = pal[1], face = "bold"),
      axis.line.y.right = element_line(color = pal[2]),
      axis.text.y.right = element_text(color = pal[2]),
      axis.ticks.y.right = element_line(color = pal[2]),
      axis.title.y.right = element_text(color = pal[2],
                                        face = "bold", margin = margin(l = 10)),
      strip.text = element_text(face = "bold.italic", size = 12))
  return(list(dati = res_sig, freq = df_frequenze, grafico = p))
}
```

# Preparation DLNM

# Constructs a distributed lag matrix for a specific climate variable. 
# For each year, it extracts a backwards sequence of 'lag' months ending 
# at the target reference month (mese_n) to capture multi-year legacy effects.

```{r}
lag_mese <- function(df, val, mese_n=9, lag = 12) {
  mesi_ita <- c("jan", "feb", "mar", "apr", "may", "jun", 
                "jul", "aug", "sep", "oct", "nov", "dec")
  # 1. Preparazione (Ordine cronologico per estrarre correttamente)
  df_clean <- df %>%
    mutate(
      mese_str = str_split_fixed(data, "-", 2)[,1],
      anno_val = as.numeric(str_split_fixed(data, "-", 2)[,2]),
      mese_num = match(mese_str, mesi_ita),
      data_obj = make_date(year = anno_val, month = mese_num, day = 1)
    ) %>%
    arrange(data_obj)
  # 2. Identifichiamo i Settembre (punti di partenza di ogni riga)
  punti_settembre <- df_clean %>% 
    filter(month(data_obj) == mese_n) %>% 
    pull(data_obj)
  # 3. Estrazione a ritroso
  estrai_sequenza <- function(data_rif) {
    res <- df_clean %>%
      filter(data_obj <= data_rif) %>% # Prendo tutto ciò che viene prima del Settembre di riferimento
      slice_tail(n = lag) %>%     # Prendo gli ultimi n mesi
      arrange(desc(data_obj)) %>%      # Inverto l'ordine: Settembre, Agosto, Luglio...
      pull(val)
    if(length(res) == lag) return(res) else return(NULL)}
  # 4. Creazione lista e pulizia anni incompleti
  lista_risultati <- lapply(punti_settembre, estrai_sequenza)
  indici_validi <- sapply(lista_risultati, function(x) !is.null(x))
  lista_finale <- lista_risultati[indici_validi]
  date_finali <- punti_settembre[indici_validi]
  # 5. Costruzione Matrice
  matrice_finale <- do.call(rbind, lista_finale)
  # Nomi righe: l'anno del settembre scelto
  rownames(matrice_finale) <- year(date_finali)
  # Nomi colonne: mostriamo il ritardo (Lag)
  mesi_ripetuti <- rep(c(mese_n:1, 12:(mese_n+1)), length.out = lag)
  offset_anni <- floor((0:(lag - 1) + 3) / 12)
  colnames(matrice_finale) <- paste0("month_", mesi_ripetuti, "_year_", offset_anni)
  return(matrice_finale)}
```

# Calculates monthly climate anomalies (deviations from long-term monthly means) 
# for a given time series, with an optional absolute value transformation, 
# and aggregates these deviations into annual total sums.

```{r}
scostamento <- function(df, mod = FALSE){
  if (nchar(df[1,1]) != 8) {stop("La prima colonna del dataframe deve contenere le date in formato gen-1901")}
  df_due <- df 
  colnames(df_due)[1] <- "data"
  df_due <- df_due %>%
    mutate (mese = as.factor(str_sub(data, start = 1, end = 3))) %>%
    group_by (mese) %>%
    mutate(across(where(is.numeric), ~ .x - mean(.x, na.rm = TRUE),
                  .names = "dif_{.col}")) %>%
    ungroup()
  if (mod) {df_tre <- df_due %>%
    mutate(across(starts_with("dif_"), abs))
  } else {df_tre <- df_due}
  df_qtr <- df_tre %>%
    mutate(anno = as.numeric(str_sub(data, -4))) %>%
    group_by(anno) %>%
    summarise(across(starts_with("dif_"), ~ sum(.x, na.rm = TRUE),
                     .names = "sum_{.col}")) %>%
    ungroup()
  return(df_qtr)}
```

# Prepares and aligns tree-ring growth data and meteorological time series 
# for mixed-effects DLNM modeling. It filters stands by climate/species group, 
# computes climate lag matrices and control variable deviations, and merges them 
# into a single unified dataset for stands with at least 10 common years.

```{r}
prepara_dlnm_misto <- function(df_tax, var_meteo, var_ctrl, month_n,
                               g_clima, g_specie, lag_n = 24) {
  raga <- df_tax
  colnames(raga)[c(3,4)] <- c("gruppo_climatico","gruppo_specie")
  if (!is.null(g_clima)) {raga <- raga %>% filter(gruppo_climatico == g_clima)}
  if (!is.null(g_specie)) {raga <- raga %>% filter(gruppo_specie == g_specie)}
  stand_selezionati <- raga %>% select(stand_id, pixel_id)
  big_df <- data.frame()
  big_X_meteo <- matrix(nrow = 0, ncol = lag_n)
  for(i in 1:nrow(stand_selezionati)) {
    s_id <- stand_selezionati$stand_id[i]
    p_id <- stand_selezionati$pixel_id[i]
    y_stand <- crono[[s_id]]
    df_meteo <- clima[[p_id]]
    if(var_meteo == "bal") {df_meteo$bal <- df_meteo$pre - df_meteo$pet}
    X_mat <- lag_mese(df_meteo, val = var_meteo, mese_n = month_n,lag = lag_n)
    df_per_scostamento <- df_meteo[, c("data", var_ctrl)]
    ctrl_annuale <- as.data.frame(scostamento(df_per_scostamento))
    colnames(ctrl_annuale) <- c("anno", var_ctrl)
    anni_comuni <- intersect(rownames(X_mat), as.character(y_stand$anno))
    anni_comuni <- intersect(anni_comuni, as.character(ctrl_annuale$anno))
    if(length(anni_comuni) > 10) {
      df_tmp <- y_stand %>% 
        filter(anno %in% anni_comuni) %>%
        left_join(ctrl_annuale %>% mutate(anno = as.numeric(anno)), by = "anno") %>%
        mutate(ID = s_id)
      X_tmp <- X_mat[anni_comuni, ]
      big_df <- rbind(big_df, df_tmp)
      big_X_meteo <- rbind(big_X_meteo, X_tmp)}}
  return(list(df = big_df, mat = big_X_meteo,
    var_ctrl_name = var_ctrl))}
```

# Adds an autoregressive term (prior year's growth, rwl_prev) to the dataset 
# by shifting tree-ring values by one year within each stand/tree ID. 
# It removes rows with missing lag values (first year of each series), 
# synchronizes the climate lag matrix accordingly, and performs an internal 
# integrity check to ensure perfect temporal alignment.

```{r}
termine_auto <- function(df, mat){
  indici_teste <- c(1, which(df$ID[-1] != df$ID[-nrow(df)]) + 1)
  rwl_slittato <- c(NA, df$rwl[-nrow(df)])
  rwl_slittato[indici_teste] <- NA
  df_aggiornato <- df
  df_aggiornato$rwl_prev <- rwl_slittato
  righe_da_tenere <- which(!is.na(df_aggiornato$rwl_prev))
  df_final <- df_aggiornato[righe_da_tenere,]
  mat_final <- mat[righe_da_tenere,]
  test_idx <- sample(2:nrow(df_final), 1)
  cor_id <- df_final$ID[test_idx]
  cor_anno <- df_final$anno[test_idx]
  valore_reale_anno_prec <- df$rwl[df$ID == cor_id & df$anno == (cor_anno - 1)]
  if(valore_reale_anno_prec == df_final$rwl_prev[test_idx]) {
    cat("✅ Allineamento Verificato!")
  } else {
    stop("❌ ERRORE DI ALLINEAMENTO!")}
  return(list(df=df_final,mat=mat_final))}
```

# Performs a triple integrity check to verify temporal consistency among 
# the original tree-ring chronology (crono), the output data frame, and the lag matrix. 
# It checks row boundaries, year ranges, potential start-year offsets (e.g., 1901–1903 
# cutoffs or rwl_prev lag adjustments), and row transitions between stands, 
# reporting alignment errors or discrepancies.

```{r}
controllo <- function(output_dlnm, report = TRUE, rwl_prev = FALSE) {
  df_out <- output_dlnm$df
  mat_out <- output_dlnm$mat
  stands_out <- unique(as.character(df_out$ID))
  n_cumulato <- 0
  cont_errori <- 0
  cont_serie <- length(stands_out)
  cat("============================================================\n")
  cat("   VERIFICA COERENZA TRIPLA: CRONO <-> DF <-> MATRICE\n")
  cat("============================================================\n\n")
  for (i in seq_along(stands_out)) {
    s_id <- stands_out[i]
    df_sub <- df_out[df_out$ID == s_id, ]
    n_righe <- nrow(df_sub)
    idx_i <- n_cumulato + 1
    idx_f <- n_cumulato + n_righe
    anno_i_out <- as.numeric(df_sub$anno[1])
    anno_f_out <- as.numeric(df_sub$anno[n_righe])
    anno_i_mat <- as.numeric(rownames(mat_out)[idx_i])
    anno_f_mat <- as.numeric(rownames(mat_out)[idx_f])
    anno_i_crono <- as.numeric(crono[[s_id]]$anno[1])
    anno_f_crono <- as.numeric(crono[[s_id]]$anno[nrow(crono[[s_id]])])
    if(report){cat(sprintf("SITO: [%s] (Righe: %d)\n", s_id, n_righe))
    cat(sprintf("  • Posizione Matrice: righe %d a %d\n", idx_i, idx_f))
    cat(sprintf("  • Range Anni Matrice: %d --> %d\n", anno_i_mat, anno_f_mat))
    cat(sprintf("  • Range Anni Dataframe: %d --> %d\n", anno_i_out, anno_f_out))
    cat(sprintf("  • Range Anni Crono:   %d --> %d\n", anno_i_crono, anno_f_crono))}
    anno_x <- ifelse(ncol(mat_out)>9 & ncol(mat_out)<22, 1902, 1903)
    cont_i_crono <- ifelse(anno_i_crono>1902, anno_i_crono, anno_x)
    if(rwl_prev) {cont_i_crono <- cont_i_crono+1}
    cont_f_crono <- anno_f_crono
    inizio_ok <- (anno_i_out == cont_i_crono && anno_i_mat == cont_i_crono)
    fine_ok <- (anno_f_out == anno_f_crono && anno_f_mat == anno_f_crono)
    if (inizio_ok && fine_ok) {
      if(report){cat("  ✅ Allineamento perfetto (o deroga 1901-03 rispettata)\n")}
    } else { cont_errori <- cont_errori+1
      if(report){cat("  ⚠️ ATTENZIONE: Discrepanza rilevata!\n")
      if(!inizio_ok) cat(sprintf("     - Errore INIZIO: Crono(%d) vs Output(%d)\n", anno_i_crono, anno_i_out))
      if(!fine_ok)   cat(sprintf("     - Errore FINE:   Crono(%d) vs Output(%d)\n", anno_f_crono, anno_f_out))}}
    if (i < length(stands_out)) {
      anno_f_attuale_mat <- rownames(mat_out)[idx_f]
      anno_i_prossimo_mat <- rownames(mat_out)[idx_f + 1]
      prossimo_id <- stands_out[i+1]
      if(report){cat(sprintf("  • GRADINO: Fine [%s]: %s | Inizio [%s]: %s\n", 
                  s_id, anno_f_attuale_mat, prossimo_id, anno_i_prossimo_mat))
      cat("    🔗 Salto temporale rilevato (corretto per cambio sito).\n")
    cat("------------------------------------------------------------\n")}}
    n_cumulato <- idx_f}
  cat(sprintf("NUMERO SITI: %s\n", cont_serie))
  cat(sprintf("NUMERO DISCREPANSE: %s\n", cont_errori))}
```

# Run the model

```{r}
get_dlnm_pred <- function(model, cb_object, cb_name = "final_cb", cen_val = 0) {
  indices <- grep(paste0("^", cb_name), names(coef(model)))
  if(length(indices) == 0) {
    stop("Errore: Non ho trovato coefficienti che corrispondono a: ", cb_name)}
  beta <- coef(model)[indices]
  V <- vcov(model)[indices, indices]
  names(beta) <- gsub(paste0("^", cb_name), "", names(beta))
  rownames(V) <- colnames(V) <- names(beta)
  pred <- crosspred(cb_object, coef = beta, vcov = V, 
                    model.link = "log", cen = cen_val)
  return(pred)}
```

# Show the charts

```{r}
confronto_lag <- function(lista_pred, valore_spei, nomi_gruppi, etichette_mesi,
                          lag_n = 24) {
  estrai_slice <- function(pred_obj, v_var, nome) {
    idx <- which.min(abs(pred_obj$predvar - v_var))
    data.frame(lag = 0:(lag_n-1),
      fit = as.vector(pred_obj$matRRfit[idx, ]),
      low = as.vector(pred_obj$matRRlow[idx, ]),
      high = as.vector(pred_obj$matRRhigh[idx, ]),
      gruppo = nome)}
  df_confronto <- mapply(estrai_slice, lista_pred, valore_spei, nomi_gruppi, SIMPLIFY = FALSE) %>%
    bind_rows()
  ggplot(df_confronto, aes(x = lag, y = fit, color = gruppo, fill = gruppo)) +
    scale_x_reverse(breaks = 0:(lag_n-1), labels = etichette_mesi[1:lag_n]) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "black") +
    geom_ribbon(aes(ymin = low, ymax = high), alpha = 0.15, color = NA) +
    geom_line(linewidth = 1) +
    labs(title = paste("SPEI = ",valore_spei," | Tree Growth Response by Genus"),
         x = "Lag (Months)", y = "Regression Coefficient", fill = "Genus", color = "Genus") +
    theme_light() + 
    theme(panel.grid.major = element_blank(), # Toglie la griglia principale
          panel.grid.minor = element_blank(),
          axis.text.x = element_text(angle = 35, hjust = 1),
          legend.position = "bottom",
          axis.title.y = element_text(margin = margin(r = 15)),
          axis.title.x = element_text(margin = margin(t = 11)),
          plot.title = element_text(size = 15),
          legend.text = element_text(face = "bold"))}
```

```{r}
grafico_unito <- function(..., nomi, mesi, titolo) {
  lista_pred <- list(...)
  nomi_modelli <- nomi
  if (is.null(nomi_modelli)) {
    nomi_modelli <- as.character(substitute(list(...))[-1])}
  prepare_data <- function(pred_obj, model_name) {
    mat <- pred_obj$matfit
    df <- expand.grid(spei = pred_obj$predvar, lag = 0:(ncol(mat) - 1))
    df$fit <- as.vector(mat) 
    df$model <- model_name
    return(df)}
  data_all <- do.call(rbind, lapply(seq_along(lista_pred), function(i) {
    prepare_data(lista_pred[[i]], nomi_modelli[i])}))
  data_all$model <- factor(data_all$model, levels = nomi_modelli)
  mia_palette <- colorRampPalette(c("firebrick", "white", "dodgerblue4"))(30)
  max_abs_global <- max(abs(data_all$fit), na.rm = TRUE)
  
  ggplot(data_all, aes(x = spei, y = lag, z = fit)) +
    stat_contour_filled(aes(fill = after_stat(level_mid)), 
                        breaks = seq(-max_abs_global, max_abs_global, length.out = 31)) +
    facet_wrap(~model, nrow = 1) + 
    scale_fill_gradientn(colors = mia_palette,
                         limits = c(-max_abs_global, max_abs_global), name = "Coefficient", 
                             guide = guide_colorbar(
      direction = "horizontal", # Forza l'orientamento orizzontale
      barwidth = 14,            # Larghezza della barra (in unità relative)
      barheight = 1.1,          # Altezza della barra (sottile!)
      title.position = "top",   # Titolo sopra la barra
      title.hjust = 0.5,        # Titolo centrato
      frame.colour = "black", 
      ticks.colour = "black"
    )) +
       scale_y_continuous(breaks = 0:(length(mesi) - 1), 
                       labels = mesi, expand = c(0,0)) +
    theme_minimal() + 
    theme(legend.position = "bottom", 
    legend.margin = margin(t = 10),
      panel.spacing.x = unit(0, "lines"),
      panel.grid.major = element_line(color = "gray80", linewidth = 0.4),
      panel.grid.minor = element_blank(),
      panel.ontop = FALSE,
      strip.text = element_text(face = "bold", size = 13, ),
      axis.title.x = element_text(size = 12, margin = margin(t = 11)),
      plot.title = element_text(size = 16),
      axis.title.y = element_text(margin = margin(r = 15)),
      axis.text.y = element_text(size = 11)) +
    labs(title = titolo, x = "SPEI", y = "Lag (Months)")}
```























