---
title: "Pacchetto"
author: "Giacomo"
date: "2026-07-21"
output: pdf_document
---

# Metadati

```{r}
leggi_tasso <- function(file_txt, vet_specie){
  file_path <- file.path("dati",file_txt)
  testo_grezzo <- readChar(file_path, file.info(file_path)$size)
  tassi_caricati <- unlist(strsplit(testo_grezzo, "\",\\s*\"|\",\""))
  tassi_caricati <- gsub("^\"|\"$", "", tassi_caricati)
  tassi_caricati <- gsub("[\r\n\t]+", " ", tassi_caricati)
  extract_taxa <- function(string, label) {
    pattern <- paste0(label, "\\s*([A-Za-z\\s\\-\\.]+?)\\s*\\(")
    match <- str_match(string, pattern)
    return(match[, 2]) }
  Tax_Final <- data.frame(
    ID_Specie = unique(vet_specie),
    Kingdom   = sapply(tassi_caricati, extract_taxa, "Kingdom"),
    Phylum    = sapply(tassi_caricati, extract_taxa, "Phylum"),
    Class     = sapply(tassi_caricati, extract_taxa, "Class"),
    Category  = sapply(tassi_caricati, extract_taxa, "Category"),
    Order     = sapply(tassi_caricati, extract_taxa, "Order"),
    Family    = sapply(tassi_caricati, extract_taxa, "Family"),
    Genus     = sapply(tassi_caricati, extract_taxa, "Genus"),
    Species = sapply(tassi_caricati, extract_taxa, "Species"))
  Tax_Final <- as.data.frame(lapply(Tax_Final, trimws))
  return(Tax_Final)}
```


```{r}
ec <- function(file_path,metadata_path){
  result <- list(metadata = list(), header = list())
  metadata <- read.delim(metadata_path)
  nome_stand_up <- toupper(str_sub(file_path, start=6, end=-5))#ATTENZIONE AL PUNTO DI START 
  if (nome_stand_up %in% metadata$ITRDB_Code) {
    coordinate <- metadata[metadata$ITRDB_Code==nome_stand_up,c("Latitude","Longitude")]
    info_stand <- metadata[metadata$ITRDB_Code==nome_stand_up,c("Elevation","Species")]
    coor <- c(coordinate$Longitude, coordinate$Latitude)
    result$metadata$coordinate <- coor
    result$metadata$info <- info_stand}
  coordinate_estratte <- ecfh(file_path,lista_specie=unique(metadata$Species))
  if (!is.null(coordinate_estratte$coord_Y) && 
      !is.null(coordinate_estratte$coord_X)){
    coor_extr <- c(coordinate_estratte$coord_Y,coordinate_estratte$coord_X)
    result$header$coordinate <- coor_extr}
  if (!is.null(coordinate_estratte$Specie)){
    result$header$specie <- coordinate_estratte$Specie}
  result$header$head <- coordinate_estratte$header
  return(result)}
```

```{r}
ecfh <- function(file_path,lista_specie) {
  result <- list()
  header_lines <- tryCatch({
    readLines(file_path, n = 3, warn = FALSE)
  }, error = function(e) {
    warning("Cannot read file: ", file_path)
    return(NULL)})
  if (is.null(header_lines)) return(result)
  header_text <- str_squish(paste(header_lines, collapse = " "))
  result$header <- header_text
  # Estrazione Specie
  tutte_quattro <- unlist(str_extract_all(header_text, "\\b[A-Z]{4}\\b"))
  if (length(tutte_quattro) == 1) result$Specie <- tutte_quattro
  match_lista <- intersect(tutte_quattro, lista_specie)
  if (length(match_lista) == 1) result$Specie <- match_lista
  # --- PATTERN 1 (M / N / E) ---
  m1 <- str_match(header_text, "(-?[[:digit:]]+)\\s*M\\s*(-?[[:digit:]]+)\\s*N?(-?[[:digit:]]+)\\s*E?")
  if (!is.na(m1[1, 1])) {
    result$coord_X <- as.numeric(m1[1, 3]) / 100
    result$coord_Y <- as.numeric(m1[1, 4]) / 100
    return(result)}
  # --- NUOVO PATTERN 4 (Formato +DDMM-DDDMM o simile) ---
  # Cerca un segno (+ o -) seguito da cifre, poi un altro segno e altre cifre
  m4 <- str_match(header_text, "([+-][[:digit:]]{3,5})([+-][[:digit:]]{3,6})")
  if (!is.na(m4[1, 1])) {
    result$coord_X <- as.numeric(m4[1, 2]) / 100
    result$coord_Y <- as.numeric(m4[1, 3]) / 100
    return(result)}
  # --- PATTERN 2 (Spazi e underscore) ---
  m2 <- str_match(header_text, "(-?[[:digit:]]+\\.?[[:digit:]]*)\\s+(-?[[:digit:]]+\\.?[[:digit:]]*)\\s+_{1,}")
  if (!is.na(m2[1, 1])) {
    val2 <- as.numeric(m2[1, 2])
    val3 <- as.numeric(m2[1, 3])
    if (val2 < 1000) {
      result$coord_X <- val3
      result$coord_Y <- val2
    } else {
      result$coord_X <- val2 / 100
      result$coord_Y <- val3 / 100}
    return(result)}
  # --- PATTERN 3 (N / W) ---
  m3 <- str_match(header_text, "\\s*(-?[[:digit:]]+)\\s*N(-?[[:digit:]]+)\\s*W")
  if (!is.na(m3[1, 1])) {
    result$coord_X <- as.numeric(m3[1, 1]) / 100
    result$coord_Y <- as.numeric(m3[1, 2]) / 100
    return(result)}
  # --- PATTERN 10 (Fisso 10 caratteri) ---
  m10 <- str_match(header_text, "(.{10})\\s+_")
  if (!is.na(m10[1, 1])) {
    raw_val <- m10[1, 2]
    result$coord_X <- as.numeric(substr(raw_val, 1, 5)) / 100
    result$coord_Y <- as.numeric(substr(raw_val, 6, 10)) / 100
    return(result)}
  return(result)}
```

# Object clima creation

```{r}
estrazione_clima <- function(df_unici) {
  file_clima <- c("spei01.nc","CRU_TS_tmp.nc","CRU_TS_tmn.nc",
                  "CRU_TS_pre.nc","CRU_TS_pet.nc")
  nomi <- c("spei","tmp","tmn","pre","pet")
  lista_meteo_raw <- list()
  for(i in 1:nrow(df_unici)) {
    p_id <- df_unici$pixel_id[i]
    coor <- c(df_unici$longitude[i],df_unici$latitude[i])
    df_pixel_accumulato <- NULL
    for(j in 1:length(file_clima)) {
      per <- file.path("dati",file_clima[j])
      df_temp <- extraction(coor, per, nome = nomi[j])
      if (is.null(df_pixel_accumulato)) {df_pixel_accumulato <- df_temp
      } else {
        df_pixel_accumulato <- merge(df_pixel_accumulato, df_temp, 
                                     by = "data", all = TRUE)}}
    lista_meteo_raw[[p_id]] <- df_pixel_accumulato
    if(i %% 50 == 0) cat("Processati", i, "pixel su", nrow(df_unici), "\n")}
  return(lista_meteo_raw)}
```


# Preparation DLNM

# Constructs a distributed lag matrix for a specific climate variable. 
# For each year, it extracts a backwards sequence of 'lag' months ending 
# at the target reference month (mese_n) to capture multi-year legacy effects.

```{r}
lag_mese <- function(df, val, mese_n=9, lag = 12) {
  mesi_ita <- c("jan", "feb", "mar", "apr", "may", "jun", 
                "jul", "aug", "sep", "oct", "nov", "dec")
  # 1. Preparazione (Ordine cronologico per estrarre correttamente)
  df_clean <- df %>%
    mutate(
      mese_str = str_split_fixed(data, "-", 2)[,1],
      anno_val = as.numeric(str_split_fixed(data, "-", 2)[,2]),
      mese_num = match(mese_str, mesi_ita),
      data_obj = make_date(year = anno_val, month = mese_num, day = 1)
    ) %>%
    arrange(data_obj)
  # 2. Identifichiamo i Settembre (punti di partenza di ogni riga)
  punti_settembre <- df_clean %>% 
    filter(month(data_obj) == mese_n) %>% 
    pull(data_obj)
  # 3. Estrazione a ritroso
  estrai_sequenza <- function(data_rif) {
    res <- df_clean %>%
      filter(data_obj <= data_rif) %>% # Prendo tutto ciò che viene prima del Settembre di riferimento
      slice_tail(n = lag) %>%     # Prendo gli ultimi n mesi
      arrange(desc(data_obj)) %>%      # Inverto l'ordine: Settembre, Agosto, Luglio...
      pull(val)
    if(length(res) == lag) return(res) else return(NULL)}
  # 4. Creazione lista e pulizia anni incompleti
  lista_risultati <- lapply(punti_settembre, estrai_sequenza)
  indici_validi <- sapply(lista_risultati, function(x) !is.null(x))
  lista_finale <- lista_risultati[indici_validi]
  date_finali <- punti_settembre[indici_validi]
  # 5. Costruzione Matrice
  matrice_finale <- do.call(rbind, lista_finale)
  # Nomi righe: l'anno del settembre scelto
  rownames(matrice_finale) <- year(date_finali)
  # Nomi colonne: mostriamo il ritardo (Lag)
  mesi_ripetuti <- rep(c(mese_n:1, 12:(mese_n+1)), length.out = lag)
  offset_anni <- floor((0:(lag - 1) + 3) / 12)
  colnames(matrice_finale) <- paste0("month_", mesi_ripetuti, "_year_", offset_anni)
  return(matrice_finale)}
```

# Calculates monthly climate anomalies (deviations from long-term monthly means) 
# for a given time series, with an optional absolute value transformation, 
# and aggregates these deviations into annual total sums.

```{r}
scostamento <- function(df, mod = FALSE){
  if (nchar(df[1,1]) != 8) {stop("La prima colonna del dataframe deve contenere le date in formato gen-1901")}
  df_due <- df 
  colnames(df_due)[1] <- "data"
  df_due <- df_due %>%
    mutate (mese = as.factor(str_sub(data, start = 1, end = 3))) %>%
    group_by (mese) %>%
    mutate(across(where(is.numeric), ~ .x - mean(.x, na.rm = TRUE),
                  .names = "dif_{.col}")) %>%
    ungroup()
  if (mod) {df_tre <- df_due %>%
    mutate(across(starts_with("dif_"), abs))
  } else {df_tre <- df_due}
  df_qtr <- df_tre %>%
    mutate(anno = as.numeric(str_sub(data, -4))) %>%
    group_by(anno) %>%
    summarise(across(starts_with("dif_"), ~ sum(.x, na.rm = TRUE),
                     .names = "sum_{.col}")) %>%
    ungroup()
  return(df_qtr)}
```

# Prepares and aligns tree-ring growth data and meteorological time series 
# for mixed-effects DLNM modeling. It filters stands by climate/species group, 
# computes climate lag matrices and control variable deviations, and merges them 
# into a single unified dataset for stands with at least 10 common years.

```{r}
prepara_dlnm_misto <- function(df_tax, var_meteo, var_ctrl, month_n,
                               g_clima, g_specie, lag_n = 24) {
  raga <- df_tax
  colnames(raga)[c(3,4)] <- c("gruppo_climatico","gruppo_specie")
  if (!is.null(g_clima)) {raga <- raga %>% filter(gruppo_climatico == g_clima)}
  if (!is.null(g_specie)) {raga <- raga %>% filter(gruppo_specie == g_specie)}
  stand_selezionati <- raga %>% select(stand_id, pixel_id)
  big_df <- data.frame()
  big_X_meteo <- matrix(nrow = 0, ncol = lag_n)
  for(i in 1:nrow(stand_selezionati)) {
    s_id <- stand_selezionati$stand_id[i]
    p_id <- stand_selezionati$pixel_id[i]
    y_stand <- crono[[s_id]]
    df_meteo <- clima[[p_id]]
    if(var_meteo == "bal") {df_meteo$bal <- df_meteo$pre - df_meteo$pet}
    X_mat <- lag_mese(df_meteo, val = var_meteo, mese_n = month_n,lag = lag_n)
    df_per_scostamento <- df_meteo[, c("data", var_ctrl)]
    ctrl_annuale <- as.data.frame(scostamento(df_per_scostamento))
    colnames(ctrl_annuale) <- c("anno", var_ctrl)
    anni_comuni <- intersect(rownames(X_mat), as.character(y_stand$anno))
    anni_comuni <- intersect(anni_comuni, as.character(ctrl_annuale$anno))
    if(length(anni_comuni) > 10) {
      df_tmp <- y_stand %>% 
        filter(anno %in% anni_comuni) %>%
        left_join(ctrl_annuale %>% mutate(anno = as.numeric(anno)), by = "anno") %>%
        mutate(ID = s_id)
      X_tmp <- X_mat[anni_comuni, ]
      big_df <- rbind(big_df, df_tmp)
      big_X_meteo <- rbind(big_X_meteo, X_tmp)}}
  return(list(df = big_df, mat = big_X_meteo,
    var_ctrl_name = var_ctrl))}
```

# Adds an autoregressive term (prior year's growth, rwl_prev) to the dataset 
# by shifting tree-ring values by one year within each stand/tree ID. 
# It removes rows with missing lag values (first year of each series), 
# synchronizes the climate lag matrix accordingly, and performs an internal 
# integrity check to ensure perfect temporal alignment.

```{r}
termine_auto <- function(df, mat){
  indici_teste <- c(1, which(df$ID[-1] != df$ID[-nrow(df)]) + 1)
  rwl_slittato <- c(NA, df$rwl[-nrow(df)])
  rwl_slittato[indici_teste] <- NA
  df_aggiornato <- df
  df_aggiornato$rwl_prev <- rwl_slittato
  righe_da_tenere <- which(!is.na(df_aggiornato$rwl_prev))
  df_final <- df_aggiornato[righe_da_tenere,]
  mat_final <- mat[righe_da_tenere,]
  test_idx <- sample(2:nrow(df_final), 1)
  cor_id <- df_final$ID[test_idx]
  cor_anno <- df_final$anno[test_idx]
  valore_reale_anno_prec <- df$rwl[df$ID == cor_id & df$anno == (cor_anno - 1)]
  if(valore_reale_anno_prec == df_final$rwl_prev[test_idx]) {
    cat("✅ Allineamento Verificato!")
  } else {
    stop("❌ ERRORE DI ALLINEAMENTO!")}
  return(list(df=df_final,mat=mat_final))}
```

# Performs a triple integrity check to verify temporal consistency among 
# the original tree-ring chronology (crono), the output data frame, and the lag matrix. 
# It checks row boundaries, year ranges, potential start-year offsets (e.g., 1901–1903 
# cutoffs or rwl_prev lag adjustments), and row transitions between stands, 
# reporting alignment errors or discrepancies.

```{r}
controllo <- function(output_dlnm, report = TRUE, rwl_prev = FALSE) {
  df_out <- output_dlnm$df
  mat_out <- output_dlnm$mat
  stands_out <- unique(as.character(df_out$ID))
  n_cumulato <- 0
  cont_errori <- 0
  cont_serie <- length(stands_out)
  cat("============================================================\n")
  cat("   VERIFICA COERENZA TRIPLA: CRONO <-> DF <-> MATRICE\n")
  cat("============================================================\n\n")
  for (i in seq_along(stands_out)) {
    s_id <- stands_out[i]
    df_sub <- df_out[df_out$ID == s_id, ]
    n_righe <- nrow(df_sub)
    idx_i <- n_cumulato + 1
    idx_f <- n_cumulato + n_righe
    anno_i_out <- as.numeric(df_sub$anno[1])
    anno_f_out <- as.numeric(df_sub$anno[n_righe])
    anno_i_mat <- as.numeric(rownames(mat_out)[idx_i])
    anno_f_mat <- as.numeric(rownames(mat_out)[idx_f])
    anno_i_crono <- as.numeric(crono[[s_id]]$anno[1])
    anno_f_crono <- as.numeric(crono[[s_id]]$anno[nrow(crono[[s_id]])])
    if(report){cat(sprintf("SITO: [%s] (Righe: %d)\n", s_id, n_righe))
    cat(sprintf("  • Posizione Matrice: righe %d a %d\n", idx_i, idx_f))
    cat(sprintf("  • Range Anni Matrice: %d --> %d\n", anno_i_mat, anno_f_mat))
    cat(sprintf("  • Range Anni Dataframe: %d --> %d\n", anno_i_out, anno_f_out))
    cat(sprintf("  • Range Anni Crono:   %d --> %d\n", anno_i_crono, anno_f_crono))}
    anno_x <- ifelse(ncol(mat_out)>9 & ncol(mat_out)<22, 1902, 1903)
    cont_i_crono <- ifelse(anno_i_crono>1902, anno_i_crono, anno_x)
    if(rwl_prev) {cont_i_crono <- cont_i_crono+1}
    cont_f_crono <- anno_f_crono
    inizio_ok <- (anno_i_out == cont_i_crono && anno_i_mat == cont_i_crono)
    fine_ok <- (anno_f_out == anno_f_crono && anno_f_mat == anno_f_crono)
    if (inizio_ok && fine_ok) {
      if(report){cat("  ✅ Allineamento perfetto (o deroga 1901-03 rispettata)\n")}
    } else { cont_errori <- cont_errori+1
      if(report){cat("  ⚠️ ATTENZIONE: Discrepanza rilevata!\n")
      if(!inizio_ok) cat(sprintf("     - Errore INIZIO: Crono(%d) vs Output(%d)\n", anno_i_crono, anno_i_out))
      if(!fine_ok)   cat(sprintf("     - Errore FINE:   Crono(%d) vs Output(%d)\n", anno_f_crono, anno_f_out))}}
    if (i < length(stands_out)) {
      anno_f_attuale_mat <- rownames(mat_out)[idx_f]
      anno_i_prossimo_mat <- rownames(mat_out)[idx_f + 1]
      prossimo_id <- stands_out[i+1]
      if(report){cat(sprintf("  • GRADINO: Fine [%s]: %s | Inizio [%s]: %s\n", 
                  s_id, anno_f_attuale_mat, prossimo_id, anno_i_prossimo_mat))
      cat("    🔗 Salto temporale rilevato (corretto per cambio sito).\n")
    cat("------------------------------------------------------------\n")}}
    n_cumulato <- idx_f}
  cat(sprintf("NUMERO SITI: %s\n", cont_serie))
  cat(sprintf("NUMERO DISCREPANSE: %s\n", cont_errori))}
```

# Run the model

```{r}
get_dlnm_pred <- function(model, cb_object, cb_name = "final_cb", cen_val = 0) {
  indices <- grep(paste0("^", cb_name), names(coef(model)))
  if(length(indices) == 0) {
    stop("Errore: Non ho trovato coefficienti che corrispondono a: ", cb_name)}
  beta <- coef(model)[indices]
  V <- vcov(model)[indices, indices]
  names(beta) <- gsub(paste0("^", cb_name), "", names(beta))
  rownames(V) <- colnames(V) <- names(beta)
  pred <- crosspred(cb_object, coef = beta, vcov = V, 
                    model.link = "log", cen = cen_val)
  return(pred)}
```

# Show the charts

```{r}
confronto_lag <- function(lista_pred, valore_spei, nomi_gruppi, etichette_mesi,
                          lag_n = 24) {
  estrai_slice <- function(pred_obj, v_var, nome) {
    idx <- which.min(abs(pred_obj$predvar - v_var))
    data.frame(lag = 0:(lag_n-1),
      fit = as.vector(pred_obj$matRRfit[idx, ]),
      low = as.vector(pred_obj$matRRlow[idx, ]),
      high = as.vector(pred_obj$matRRhigh[idx, ]),
      gruppo = nome)}
  df_confronto <- mapply(estrai_slice, lista_pred, valore_spei, nomi_gruppi, SIMPLIFY = FALSE) %>%
    bind_rows()
  ggplot(df_confronto, aes(x = lag, y = fit, color = gruppo, fill = gruppo)) +
    scale_x_reverse(breaks = 0:(lag_n-1), labels = etichette_mesi[1:lag_n]) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "black") +
    geom_ribbon(aes(ymin = low, ymax = high), alpha = 0.15, color = NA) +
    geom_line(linewidth = 1) +
    labs(title = paste("SPEI = ",valore_spei," | Tree Growth Response by Genus"),
         x = "Lag (Months)", y = "Regression Coefficient", fill = "Genus", color = "Genus") +
    theme_light() + 
    theme(panel.grid.major = element_blank(), # Toglie la griglia principale
          panel.grid.minor = element_blank(),
          axis.text.x = element_text(angle = 35, hjust = 1),
          legend.position = "bottom",
          axis.title.y = element_text(margin = margin(r = 15)),
          axis.title.x = element_text(margin = margin(t = 11)),
          plot.title = element_text(size = 15),
          legend.text = element_text(face = "bold"))}
```

```{r}
grafico_unito <- function(..., nomi, mesi, titolo) {
  lista_pred <- list(...)
  nomi_modelli <- nomi
  if (is.null(nomi_modelli)) {
    nomi_modelli <- as.character(substitute(list(...))[-1])}
  prepare_data <- function(pred_obj, model_name) {
    mat <- pred_obj$matfit
    df <- expand.grid(spei = pred_obj$predvar, lag = 0:(ncol(mat) - 1))
    df$fit <- as.vector(mat) 
    df$model <- model_name
    return(df)}
  data_all <- do.call(rbind, lapply(seq_along(lista_pred), function(i) {
    prepare_data(lista_pred[[i]], nomi_modelli[i])}))
  data_all$model <- factor(data_all$model, levels = nomi_modelli)
  mia_palette <- colorRampPalette(c("firebrick", "white", "dodgerblue4"))(30)
  max_abs_global <- max(abs(data_all$fit), na.rm = TRUE)
  
  ggplot(data_all, aes(x = spei, y = lag, z = fit)) +
    stat_contour_filled(aes(fill = after_stat(level_mid)), 
                        breaks = seq(-max_abs_global, max_abs_global, length.out = 31)) +
    facet_wrap(~model, nrow = 1) + 
    scale_fill_gradientn(colors = mia_palette,
                         limits = c(-max_abs_global, max_abs_global), name = "Coefficient", 
                             guide = guide_colorbar(
      direction = "horizontal", # Forza l'orientamento orizzontale
      barwidth = 14,            # Larghezza della barra (in unità relative)
      barheight = 1.1,          # Altezza della barra (sottile!)
      title.position = "top",   # Titolo sopra la barra
      title.hjust = 0.5,        # Titolo centrato
      frame.colour = "black", 
      ticks.colour = "black"
    )) +
       scale_y_continuous(breaks = 0:(length(mesi) - 1), 
                       labels = mesi, expand = c(0,0)) +
    theme_minimal() + 
    theme(legend.position = "bottom", 
    legend.margin = margin(t = 10),
      panel.spacing.x = unit(0, "lines"),
      panel.grid.major = element_line(color = "gray80", linewidth = 0.4),
      panel.grid.minor = element_blank(),
      panel.ontop = FALSE,
      strip.text = element_text(face = "bold", size = 13, ),
      axis.title.x = element_text(size = 12, margin = margin(t = 11)),
      plot.title = element_text(size = 16),
      axis.title.y = element_text(margin = margin(r = 15)),
      axis.text.y = element_text(size = 11)) +
    labs(title = titolo, x = "SPEI", y = "Lag (Months)")}
```























