---
title: "Pacchetto"
author: "Giacomo"
date: "2026-07-21"
output: pdf_document
---

# Metadati

# Preparation DLNM

# Constructs a distributed lag matrix for a specific climate variable. 
# For each year, it extracts a backwards sequence of 'lag' months ending 
# at the target reference month (mese_n) to capture multi-year legacy effects.

```{r}
lag_mese <- function(df, val, mese_n=9, lag = 12) {
  mesi_ita <- c("gen", "feb", "mar", "apr", "mag", "giu", 
                "lug", "ago", "set", "ott", "nov", "dic")
  # 1. Preparazione (Ordine cronologico per estrarre correttamente)
  df_clean <- df %>%
    mutate(
      mese_str = str_split_fixed(data, "-", 2)[,1],
      anno_val = as.numeric(str_split_fixed(data, "-", 2)[,2]),
      mese_num = match(mese_str, mesi_ita),
      data_obj = make_date(year = anno_val, month = mese_num, day = 1)
    ) %>%
    arrange(data_obj)
  # 2. Identifichiamo i Settembre (punti di partenza di ogni riga)
  punti_settembre <- df_clean %>% 
    filter(month(data_obj) == mese_n) %>% 
    pull(data_obj)
  # 3. Estrazione a ritroso
  estrai_sequenza <- function(data_rif) {
    res <- df_clean %>%
      filter(data_obj <= data_rif) %>% # Prendo tutto ciò che viene prima del Settembre di riferimento
      slice_tail(n = lag) %>%     # Prendo gli ultimi n mesi
      arrange(desc(data_obj)) %>%      # Inverto l'ordine: Settembre, Agosto, Luglio...
      pull(val)
    if(length(res) == lag) return(res) else return(NULL)}
  # 4. Creazione lista e pulizia anni incompleti
  lista_risultati <- lapply(punti_settembre, estrai_sequenza)
  indici_validi <- sapply(lista_risultati, function(x) !is.null(x))
  lista_finale <- lista_risultati[indici_validi]
  date_finali <- punti_settembre[indici_validi]
  # 5. Costruzione Matrice
  matrice_finale <- do.call(rbind, lista_finale)
  # Nomi righe: l'anno del settembre scelto
  rownames(matrice_finale) <- year(date_finali)
  # Nomi colonne: mostriamo il ritardo (Lag)
  mesi_ripetuti <- rep(c(mese_n:1, 12:(mese_n+1)), length.out = lag)
  offset_anni <- floor((0:(lag - 1) + 3) / 12)
  colnames(matrice_finale) <- paste0("mese_", mesi_ripetuti, "_anno_", offset_anni)
  return(matrice_finale)}
```

# Calculates monthly climate anomalies (deviations from long-term monthly means) 
# for a given time series, with an optional absolute value transformation, 
# and aggregates these deviations into annual total sums.

```{r}
scostamento <- function(df, mod = FALSE){
  if (nchar(df[1,1]) != 8) {stop("La prima colonna del dataframe deve contenere le date in formato gen-1901")}
  df_due <- df 
  colnames(df_due)[1] <- "data"
  df_due <- df_due %>%
    mutate (mese = as.factor(str_sub(data, start = 1, end = 3))) %>%
    group_by (mese) %>%
    mutate(across(where(is.numeric), ~ .x - mean(.x, na.rm = TRUE),
                  .names = "dif_{.col}")) %>%
    ungroup()
  if (mod) {df_tre <- df_due %>%
    mutate(across(starts_with("dif_"), abs))
  } else {df_tre <- df_due}
  df_qtr <- df_tre %>%
    mutate(anno = as.numeric(str_sub(data, -4))) %>%
    group_by(anno) %>%
    summarise(across(starts_with("dif_"), ~ sum(.x, na.rm = TRUE),
                     .names = "sum_{.col}")) %>%
    ungroup()
  return(df_qtr)}
```

# Prepares and aligns tree-ring growth data and meteorological time series 
# for mixed-effects DLNM modeling. It filters stands by climate/species group, 
# computes climate lag matrices and control variable deviations, and merges them 
# into a single unified dataset for stands with at least 10 common years.

```{r}
prepara_dlnm_misto <- function(df_tax, var_meteo, var_ctrl, month_n,
                               g_clima, g_specie, lag_n = 24) {
  raga <- df_tax
  colnames(raga)[c(3,4)] <- c("gruppo_climatico","gruppo_specie")
  if (!is.null(g_clima)) {raga <- raga %>% filter(gruppo_climatico == g_clima)}
  if (!is.null(g_specie)) {raga <- raga %>% filter(gruppo_specie == g_specie)}
  stand_selezionati <- raga %>% select(stand_id, pixel_id)
  big_df <- data.frame()
  big_X_meteo <- matrix(nrow = 0, ncol = lag_n)
  for(i in 1:nrow(stand_selezionati)) {
    s_id <- stand_selezionati$stand_id[i]
    p_id <- stand_selezionati$pixel_id[i]
    y_stand <- crono[[s_id]]
    df_meteo <- clima[[p_id]]
    if(var_meteo == "bal") {df_meteo$bal <- df_meteo$pre - df_meteo$pet}
    X_mat <- lag_mese(df_meteo, val = var_meteo, mese_n = month_n,lag = lag_n)
    df_per_scostamento <- df_meteo[, c("data", var_ctrl)]
    ctrl_annuale <- as.data.frame(scostamento(df_per_scostamento))
    colnames(ctrl_annuale) <- c("anno", var_ctrl)
    anni_comuni <- intersect(rownames(X_mat), as.character(y_stand$anno))
    anni_comuni <- intersect(anni_comuni, as.character(ctrl_annuale$anno))
    if(length(anni_comuni) > 10) {
      df_tmp <- y_stand %>% 
        filter(anno %in% anni_comuni) %>%
        left_join(ctrl_annuale %>% mutate(anno = as.numeric(anno)), by = "anno") %>%
        mutate(ID = s_id)
      X_tmp <- X_mat[anni_comuni, ]
      big_df <- rbind(big_df, df_tmp)
      big_X_meteo <- rbind(big_X_meteo, X_tmp)}}
  return(list(df = big_df, mat = big_X_meteo,
    var_ctrl_name = var_ctrl))}
```

# Adds an autoregressive term (prior year's growth, rwl_prev) to the dataset 
# by shifting tree-ring values by one year within each stand/tree ID. 
# It removes rows with missing lag values (first year of each series), 
# synchronizes the climate lag matrix accordingly, and performs an internal 
# integrity check to ensure perfect temporal alignment.

```{r}
termine_auto <- function(df, mat){
  indici_teste <- c(1, which(df$ID[-1] != df$ID[-nrow(df)]) + 1)
  rwl_slittato <- c(NA, df$rwl[-nrow(df)])
  rwl_slittato[indici_teste] <- NA
  df_aggiornato <- df
  df_aggiornato$rwl_prev <- rwl_slittato
  righe_da_tenere <- which(!is.na(df_aggiornato$rwl_prev))
  df_final <- df_aggiornato[righe_da_tenere,]
  mat_final <- mat[righe_da_tenere,]
  test_idx <- sample(2:nrow(df_final), 1)
  cor_id <- df_final$ID[test_idx]
  cor_anno <- df_final$anno[test_idx]
  valore_reale_anno_prec <- df$rwl[df$ID == cor_id & df$anno == (cor_anno - 1)]
  if(valore_reale_anno_prec == df_final$rwl_prev[test_idx]) {
    cat("✅ Allineamento Verificato!")
  } else {
    stop("❌ ERRORE DI ALLINEAMENTO!")}
  return(list(df=df_final,mat=mat_final))}
```

# Performs a triple integrity check to verify temporal consistency among 
# the original tree-ring chronology (crono), the output data frame, and the lag matrix. 
# It checks row boundaries, year ranges, potential start-year offsets (e.g., 1901–1903 
# cutoffs or rwl_prev lag adjustments), and row transitions between stands, 
# reporting alignment errors or discrepancies.

```{r}
controllo <- function(output_dlnm, report = TRUE, rwl_prev = FALSE) {
  df_out <- output_dlnm$df
  mat_out <- output_dlnm$mat
  stands_out <- unique(as.character(df_out$ID))
  n_cumulato <- 0
  cont_errori <- 0
  cont_serie <- length(stands_out)
  cat("============================================================\n")
  cat("   VERIFICA COERENZA TRIPLA: CRONO <-> DF <-> MATRICE\n")
  cat("============================================================\n\n")
  for (i in seq_along(stands_out)) {
    s_id <- stands_out[i]
    df_sub <- df_out[df_out$ID == s_id, ]
    n_righe <- nrow(df_sub)
    idx_i <- n_cumulato + 1
    idx_f <- n_cumulato + n_righe
    anno_i_out <- as.numeric(df_sub$anno[1])
    anno_f_out <- as.numeric(df_sub$anno[n_righe])
    anno_i_mat <- as.numeric(rownames(mat_out)[idx_i])
    anno_f_mat <- as.numeric(rownames(mat_out)[idx_f])
    anno_i_crono <- as.numeric(crono[[s_id]]$anno[1])
    anno_f_crono <- as.numeric(crono[[s_id]]$anno[nrow(crono[[s_id]])])
    if(report){cat(sprintf("SITO: [%s] (Righe: %d)\n", s_id, n_righe))
    cat(sprintf("  • Posizione Matrice: righe %d a %d\n", idx_i, idx_f))
    cat(sprintf("  • Range Anni Matrice: %d --> %d\n", anno_i_mat, anno_f_mat))
    cat(sprintf("  • Range Anni Dataframe: %d --> %d\n", anno_i_out, anno_f_out))
    cat(sprintf("  • Range Anni Crono:   %d --> %d\n", anno_i_crono, anno_f_crono))}
    anno_x <- ifelse(ncol(mat_out)>9 & ncol(mat_out)<22, 1902, 1903)
    cont_i_crono <- ifelse(anno_i_crono>1902, anno_i_crono, anno_x)
    if(rwl_prev) {cont_i_crono <- cont_i_crono+1}
    cont_f_crono <- anno_f_crono
    inizio_ok <- (anno_i_out == cont_i_crono && anno_i_mat == cont_i_crono)
    fine_ok <- (anno_f_out == anno_f_crono && anno_f_mat == anno_f_crono)
    if (inizio_ok && fine_ok) {
      if(report){cat("  ✅ Allineamento perfetto (o deroga 1901-03 rispettata)\n")}
    } else { cont_errori <- cont_errori+1
      if(report){cat("  ⚠️ ATTENZIONE: Discrepanza rilevata!\n")
      if(!inizio_ok) cat(sprintf("     - Errore INIZIO: Crono(%d) vs Output(%d)\n", anno_i_crono, anno_i_out))
      if(!fine_ok)   cat(sprintf("     - Errore FINE:   Crono(%d) vs Output(%d)\n", anno_f_crono, anno_f_out))}}
    if (i < length(stands_out)) {
      anno_f_attuale_mat <- rownames(mat_out)[idx_f]
      anno_i_prossimo_mat <- rownames(mat_out)[idx_f + 1]
      prossimo_id <- stands_out[i+1]
      if(report){cat(sprintf("  • GRADINO: Fine [%s]: %s | Inizio [%s]: %s\n", 
                  s_id, anno_f_attuale_mat, prossimo_id, anno_i_prossimo_mat))
      cat("    🔗 Salto temporale rilevato (corretto per cambio sito).\n")
    cat("------------------------------------------------------------\n")}}
    n_cumulato <- idx_f}
  cat(sprintf("NUMERO SITI: %s\n", cont_serie))
  cat(sprintf("NUMERO DISCREPANSE: %s\n", cont_errori))}
```

# Run the model

# Show the charts

























