---
title: "Answer"
author: "Giacomo Colasurdo"
output: pdf_document
---

# download rwl series

```{r include=FALSE}
library(httr)
library(rvest)
library(stringr)
library(magrittr)
```

```{r include=FALSE}
url = "https://www.ncei.noaa.gov/pub/data/paleo/treering/measurements/europe/"
# select `download = TRUE` to download file in folder: File
# Archetype is the list of name files.
archetype <- download_files(url,download = FALSE)
```

# Sample depth

```{r include=FALSE}
library(dplR)
library(ggplot2)
library(dplyr)
library(tidyr)
```

```{r}
stand_ids_disponibili <- list.files("File/")

# Genera il dataframe cumulativo su tutti i siti
df_long_finale <- prepara_df_disponibilita(
  stand_ids  = stand_ids_disponibili, 
  folder     = "File/",
  start_year = 1900, 
  end_year   = 2024
)
```

```{r}
p_cumulativo <- ggplot(df_long_finale, aes(x = Anno, y = Frequenza_Relativa, color = Tipo_Soglia)) +
  geom_line(linewidth = 1.3) + 
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10), labels = function(x) paste0(x, "%")) +
  scale_color_manual(
    values = c("Soglia_6_Alberi" = "lightsalmon", "Common_Interval_100" = "steelblue"),
    labels = c("100% Series (Common Interval)", "Min. 6 Trees (Robust Threshold)")
  ) +
  labs(
    title = "Relative Frequency of Site Availability",
    subtitle = paste("Based on", length(stand_ids_disponibili), "valid sites"),
    x = "Year", y = "Percentage of Total Sites", color = "Inclusion Criteria"
  ) +
  theme_bw() + 
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5, face = "bold"))

print(p_cumulativo)
ggsave("DendroCrhAn/sample_death_cumulative.pdf", plot = p_cumulativo, width = 9, height = 6, dpi = 300)
```


# Stand Processing

dati/ITRDBmetadata12January2022.txt

```{r}
library(dplR)
library(stringr)
```

```{r}
crono <- elabora_cronologie_rwl(dir_file = "File")
saveRDS(crono, "dati/cronologie_tutti_stand.rds")
```

# Metadata

```{r}
library(terra)
```

```{r}
output_spaziale <- estrai_metadati_e_master(
  stand_validi  = names(crono),
  path_metadata = "dati/ITRDBmetadata12January2022.txt",
  dir_file      = "File")
df_master_grezzo <- output_spaziale$df_master
punti_geografici_vect <- output_spaziale$punti_vect
```

# Coordinate solution

```{r}
audit_completo <- gestisci_punti_in_acqua(df_master_grezzo)
```

Supplementary Material for the "Coordinate Solution" Chapter of the Exploration_checks.rmd File

```{r}
audit_completo[audit_completo$stand_id=="fran10.rwl",c("lon_snap","lat_snap")] <- c(4.483,52.13)
correction <- c(rep(TRUE,5),rep(FALSE,7))
names(correction) <- audit_completo$stand_id
```

```{r}
risultato_finale <- applica_correzioni_coordinate(
  df_audit_totale = df_master_grezzo,
  df_audit_acqua  = audit_completo,
  vec_inversione  = correction)

punti_geografici_corretti <- risultato_finale$vect
df_master <- risultato_finale$audit

saveRDS(punti_geografici_corretti, "dati/punti_geografici_corretti.rds")
saveRDS(df_master, "dati/tabella_master_ITRDB_corretta.rds")
```

# Climatic zone 

```{r}
library(dplyr)
```

Group BS: arid climates
Group C is the union of
Group Cf: temperate climates and Group Cs: Mediterranean climates
Group D: cold continental climates
Group E: polar climates

```{r}
kg_map <- rast("KGC/Map_KG-Global/KG_1986-2010.grd")
clima_valori <- extract(kg_map, df_master[, c("longitude", "latitude")])
meta <- cbind(df_master, clima_valori[, c("layer")])
colnames(meta)[ncol(meta)] <- "ID"
meta <- left_join(meta, rat, by="ID")
colnames(meta)[(ncol(meta)-1):ncol(meta)] <- c("zona_id","climate_zona")
df_meta <- meta[, c("stand_id", "longitude", "latitude", "altitude",
                    "zona_id", "climate_zona", "species", "header_text")]
# An examination of the missing values reveals that they are TICO species
df_meta[is.na(df_meta$species),"species"] <- "TICO"
saveRDS(df_meta, "dati/metadata_zone_climatiche.rds")
```

# Taxonomies

More datails in Exploration_checks.Rmd File to chapter "Combine species information conifers or broadleaves"

```{r include=FALSE}
library(terra)
library(dplyr)
library(stringr)
```

```{r}
df_meta <- readRDS("dati/metadata_zone_climatiche.rds")
```

```{r}
Tax <- leggi_tasso(file_txt = "dati/Tasso.txt", df_meta$species)
# Replace missing species names with the corresponding Genus name
Tax[is.na(Tax$Species),"Species"] <- Tax[is.na(Tax$Species),"Genus"]
saveRDS(Tax, "dati/tassonomia.rds")
```

```{r}
# Join and Reclassification
tax_prep <- Tax %>% 
  rename(species = ID_Specie, Species_name = Species)

veg_prep <- left_join(df_meta, tax_prep, by = "species")
Genus_target <- c("Pinus","Quercus","Fagus")
veg <- veg_prep %>% 
  mutate(
    gruppo_climatico = case_when(
      grepl("^BS", climate_zona) ~ "B",
      grepl("^C", climate_zona) ~ "C",
      grepl("^D",  climate_zona) ~ "D",
      grepl("^ET", climate_zona) ~ "E")) %>%
  filter(Genus %in% Genus_target) %>%
  rename(gruppo_specie = Genus)

# Final Saving of meta_essential
df_veg <- veg[, c("stand_id", "gruppo_climatico", "gruppo_specie", "longitude", "latitude")]
saveRDS(df_veg, "dati/meta_group.rds")
```

# extraction of pixel

 - File esterni
meta_group.rds
CRU_TS_pre.nc

```{r}
library(ncdf4)
```

```{r}
df_veg <- readRDS("dati/meta_group.rds")
nc_data <- nc_open("dati/CRU_TS_pre.nc")
map_lon <- ncvar_get(nc_data, "lon")
map_lat <- ncvar_get(nc_data, "lat")
nc_close(nc_data)
```

```{r}
find_pixel_index <- function(lon, lat, map_lon, map_lat) {
  idx_lon <- which.min(abs(map_lon - lon))
  idx_lat <- which.min(abs(map_lat - lat))
  return(paste(idx_lon, idx_lat, sep = "_"))}
```

```{r}
df_veg$pixel_id <- mapply(find_pixel_index, 
                          df_veg$longitude, 
                          df_veg$latitude, 
                          MoreArgs = list(map_lon = map_lon, map_lat = map_lat))
saveRDS(df_veg, "dati/meta_essential.rds")
```

# Object clima creation

 - Oggetti esterni
df_tax — Dataframe di partenza che deve contenere almeno le colonne: pixel_id, longitude e latitude
 - File esterni
spei01.nc
CRU_TS_tmp.nc
CRU_TS_tmn.nc
CRU_TS_pre.nc
CRU_TS_pet.nc

```{r}
library(dplyr)
library(terra)
```

```{r}
df_veg <- readRDS("dati/meta_essential.rds")
df_pixel <- df_veg %>%
  distinct(pixel_id, .keep_all = TRUE) %>%
  select(pixel_id, longitude, latitude)
clima <- estrazione_clima(df_pixel)
saveRDS(clima, "dati/lista_clima.Rds")
```

# Cross Correlation

```{r}
correlazione_significativa_lag(df_veg,sel_orders = "Quercus")
correlazione_significativa_lag(df_veg,sel_orders = "Pinus")
correlazione_significativa_lag(df_veg,sel_orders = "Fagus")
```


# Preparation DLNM

```{r}
library(dplyr)
library(stringr)
library(lubridate)
```

```{r}
df_veg <- readRDS("dati/meta_essential.rds")
```


```{r}
dlnm_spei_fagus <- prepara_dlnm_misto(df_veg, var_meteo = "spei", var_ctrl = "tmn",
                                     month_n = 9, g_clima = NULL,
                                     g_specie = "Fagus", lag_n = 21)
dlnm_spei_quercus <- prepara_dlnm_misto(df_veg, var_meteo = "spei", var_ctrl = "tmn",
                                     month_n = 9, g_clima = NULL,
                                     g_specie = "Fagus", lag_n = 21)
dlnm_spei_pinus <- prepara_dlnm_misto(df_veg, var_meteo = "spei", var_ctrl = "tmn",
                                     month_n = 9, g_clima = NULL,
                                     g_specie = "Fagus", lag_n = 21)
```

```{r}
saveRDS(dlnm_spei_fagus,"dati/dlnm_21spei_all_tmn_Fagus.rds")
saveRDS(dlnm_spei_quercus,"dati/dlnm_21spei_all_tmn_Fagus.rds")
saveRDS(dlnm_spei_pinus,"dati/dlnm_21spei_all_tmn_Fagus.rds")
```

```{r}
faggio <- readRDS("dati/dlnm_21spei_all_tmn_Fagus.rds")
quercia <- readRDS("dati/dlnm_21spei_all_tmn_Quercus.rds")
pino <- readRDS("dati/dlnm_21spei_all_tmn_Pinus.rds")
```

```{r}
faggio_auto <- termine_auto(faggio$df, faggio$mat)
controllo(faggio_auto, report = FALSE, rwl_prev = TRUE)
faggio_auto$df$ID <- as.factor(faggio_auto$df$ID)
quercia_auto <- termine_auto(quercia$df, quercia$mat)
controllo(quercia_auto, report = FALSE, rwl_prev = TRUE)
quercia_auto$df$ID <- as.factor(quercia_auto$df$ID)
pino_auto <- termine_auto(pino$df, pino$mat)
controllo(pino_auto, report = FALSE, rwl_prev = TRUE)
pino_auto$df$ID <- as.factor(pino_auto$df$ID)
```

# Run the model

```{r}
library(dlnm)
library(mgcv)
library(splines)
```

```{r}
cb_faggio <- crossbasis(faggio_auto$mat, lag = 20,
                       argvar = list(fun = "ns", df = 2), 
                       arglag = list(fun = "ns", df = 6))
model_faggio <- bam(rwl ~ cb_faggio + rwl_prev +
                   s(ID, bs = "re") +
                   s(anno, bs = "ps", k = 20),
                   family = Gamma(link = "log"), 
                   data = faggio_auto$df,
                   method = "fREML",
                   discrete = TRUE)
pred_faggio <- get_dlnm_pred(model_faggio, cb_faggio, cb_name = "cb_faggio")
```

```{r}
cb_quercia <- crossbasis(quercia_auto$mat, lag = 20,
                       argvar = list(fun = "ns", df = 2), 
                       arglag = list(fun = "ns", df = 6))
model_quercia <- bam(rwl ~ cb_quercia + rwl_prev +
                   s(ID, bs = "re") +
                   s(anno, bs = "ps", k = 20),
                   family = Gamma(link = "log"), 
                   data = quercia_auto$df,
                   method = "fREML",
                   discrete = TRUE)
pred_quercia <- get_dlnm_pred(model_quercia, cb_quercia, cb_name = "cb_quercia")
```

```{r}
cb_pino <- crossbasis(pino_auto$mat, lag = 20,
                       argvar = list(fun = "ns", df = 2), 
                       arglag = list(fun = "ns", df = 6))
model_pino <- bam(rwl ~ cb_pino + rwl_prev +
                   s(ID, bs = "re") +
                   s(anno, bs = "ps", k = 20),
                   family = Gamma(link = "log"), 
                   data = pino_auto$df,
                   method = "fREML",
                   discrete = TRUE)
pred_pino <- get_dlnm_pred(model_pino, cb_pino, cb_name = "cb_pino")
```

# Show the charts

```{r}
library(ggplot2)
```

```{r}
mesi_sin <- c("sep", "aug", "jul", "jun", "may", "apr", "mar", "feb", "jan") 
mesi_cos <- c("dec", "nov", "oct")
mesi_eng <- c()
for (anno in c(0:2))
{mesi_eng <- c(mesi_eng,str_replace_all(paste(mesi_sin,"-",anno), " ", ""),
                        str_replace_all(paste(mesi_cos,"-",anno+1), " ", ""))}
```

```{r}
nomi <- c("Pinus","Quercus","Fagus")
confronto_lag(lista_pred = list(pred_pino,pred_quercia,pred_faggio),
              valore_spei = 2.5, etichette_mesi = mesi_eng, lag_n=21,
              nomi_gruppi = nomi)
ggsave(filename = "gasparrini_genus_pos.pdf",
       device = "pdf",width = 7, height = 5,  units = "in",dpi = 300 )
```

```{r}
png("gasparrini_genus_neg.png", width = 7, height = 5, units = "in", res = 300)
print(confronto_lag(lista_pred = list(pred_pino,pred_quercia,pred_faggio),
              valore_spei = -2.5, etichette_mesi = mesi_eng, lag_n=21,
              nomi_gruppi = nomi))
dev.off()
```

```{r}
grafico_unito(pred_pino,pred_quercia,pred_faggio, mesi=mesi_eng, nomi=nomi,
              titolo="Lagged Effects of SPEI on Tree Growth by Genus")
ggsave(filename = "gasparrini_genus.pdf", 
       device = "pdf",width = 7, height = 10,  units = "in", dpi = 300 )
```


```{r}
png("gasparrini_genus.png", width = 7, height = 6, units = "in", res = 300)
print(grafico_unito(pred_pino,pred_quercia,pred_faggio, mesi=mesi_eng, nomi=nomi,
              titolo="Lagged Effects of SPEI on Tree Growth by Genus"))
dev.off()
```

