---
title: "Answer"
author: "Giacomo Colasurdo"
output: pdf_document
---

# Preamble

## Downlaod and Load the package

```{r include=FALSE}
#install.packages("remotes")
remotes::install_github("Jackz-Juice/DendroCrhAn")
```


## Folder Creation

```{r}
folder <- "DendroCrhAn"
folder_rwl <- file.path(folder,"File")
folder_plot <- file.path(folder,"Plot")
folder_dati <- file.path(folder,"Dati")
dir.create(folder_rwl, recursive = TRUE)
dir.create(folder_plot)
dir.create(folder_dati)
```

# download rwl series

```{r include=FALSE}
library(httr)
library(rvest)
library(stringr)
library(magrittr)
```

The code downloads .rwl files containing raw European Tree Ring Width (TRW) dendrochronological series using the download_files function. The function also returns a list of the downloaded file names. Depending on your connection, the download may take some time. If the website is unreachable or you wish to skip downloading the files this way, you can access them by downloading File.rar from GitHub.
I suggest to name the folder of rwl serie `File`.

```{r include=FALSE}
url = "https://www.ncei.noaa.gov/pub/data/paleo/treering/measurements/europe/"
# select `download = TRUE` to download file in folder: `name_folder`
# Archetype is the list of name files, but it's not necessary for next part 
archetype <- download_files(url, name_folder = folder_rwl, download = TRUE)
```

# Stand Processing and Sample depth

## Stand processing

```{r include=FALSE}
library(dplR)
library(dplyr)
library(tidyr)
```

The elabora_cronologie function reads each dendrochronological series contained in every downloaded file. The series within each stand are detrended, cross-dated to determine the minimum number of annual observations, and finally compiled into the Master Chronology for the field. This is returned by the cronologie object, whereas the df_disponibilita output consists of a dataframe useful for constructing the sample depth plot.

```{r include=FALSE}
cronologie_rwl <- elabora_cronologie(dir_file = folder_rwl)
saveRDS(cronologie_rwl, file.path(folder_dati,"chronology_analysis_result.rds"))
# We need also to save separately the output related to Master Chronology Series
# Infact the object and name `crono` are essential and used until the end
crono <- cronologie_rwl$cronologie
saveRDS(crono, file.path(folder_dati,"chronology_all_stand.rds"))
df_long <- cronologie_rwl$df_disponibilita
stand_disponibili <- cronologie_rwl$stand_read
```

## Sample depth

```{r include=FALSE}
library(ggplot2)
```

```{r}
sample_depth_cum <- ggplot(df_long, aes(x = Year, y = Relative_Frequency,
                                           color = threshold_Type)) +
  geom_line(linewidth = 1.3) + 
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10),
                     labels = function(x) paste0(x, "%")) +
  scale_color_manual(
    values = c("Threshold_6_Tree" = "lightsalmon",
               "Common_Interval_100" = "steelblue"),
    labels = c("100% Series (Common Interval)",
               "Min. 6 Trees (Robust Threshold)")) +
  labs(title = "Relative Frequency of Site Availability",
       subtitle = paste("Based on",stand_disponibili, "valid sites"),
    x = "Year", y = "Percentage of Total Sites", color = "Inclusion Criteria") +
  theme_bw() + 
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5,
                                                              face = "bold"))
print(sample_depth_cum)
ggsave(file.path(folder_plot,"sample_death_cumulative.pdf"),
                 plot = sample_depth_cum, width = 9, height = 6, dpi = 300)
```

# Metadata enhancement

## Coordinate solution

```{r include=FALSE}
library(terra)
```

```{r}
output_spaziale <- estrai_metadati_e_master(
  stand_validi  = names(crono), 
  path_metadata = file.path(folder_dati,"ITRDBmetadata12January2022.txt"), 
  dir_file      = folder_rwl)
punti_vect <- output_spaziale$punti_vect
df_report  <- output_spaziale$df_coor
```


```{r}
path_raster <- file.path(folder_dati,"spei01.nc")
report_water <- gestisci_punti_in_acqua(output_spaziale,
                                        path_raster_cru = path_raster)
```

The in-depth analysis of the coordinates, along with the analysis of the water stands based on the climate maps related to the SPEI index and the rationale for the upcoming steps, are reported in the `Exploration.Rmd` file.

```{r}
correction <- data.frame(
  stand_id = c("esp098", "esp099", "esp101", "fra059", "swe359", 
               "swe360", "swe361", "swe362", "swed324"),
  correction = c(rep(FALSE, 3), rep(TRUE, 6)))

# Chiamata alla nuova funzione
output_spaziale_corr <- applica_correzioni_coordinate(
  df_report = df_report,
  report_water = report_water,  
  df_correction   = correction)

punti_geografici <- output_spaziale_corr$vect
df_meta <- output_spaziale_corr$df_meta

saveRDS(punti_geografici, file.path(folder_dati,"geographical_points.rds"))
saveRDS(df_meta, file.path(folder_dati,"table_master_ITRDB.rds"))
```

## Clima extrapolation

```{r}
library(dplyr)
```

Group BS: arid climates
Group C is the union of
Group Cf: temperate climates and Group Cs: Mediterranean climates
Group D: cold continental climates
Group E: polar climates

```{r}
kg_map <- rast("KGC/Map_KG-Global/KG_1986-2010.grd")
clima_valori <- extract(kg_map, df_master[, c("longitude", "latitude")])
meta <- cbind(df_master, clima_valori[, c("layer")])
colnames(meta)[ncol(meta)] <- "ID"
meta <- left_join(meta, rat, by="ID")
colnames(meta)[(ncol(meta)-1):ncol(meta)] <- c("zona_id","climate_zona")
df_meta <- meta[, c("stand_id", "longitude", "latitude", "altitude",
                    "zona_id", "climate_zona", "species", "header_text")]
# An examination of the missing values reveals that they are TICO species
df_meta[is.na(df_meta$species),"species"] <- "TICO"
saveRDS(df_meta, "dati/metadata_zone_climatiche.rds")
```

## Taxonomies union

More datails in Exploration_checks.Rmd File to chapter "Combine species information conifers or broadleaves"

```{r include=FALSE}
library(terra)
library(dplyr)
library(stringr)
```

```{r}
df_meta <- readRDS("dati/metadata_zone_climatiche.rds")
```

```{r}
Tax <- leggi_tasso(file_txt = "dati/Tasso.txt", df_meta$species)
# Replace missing species names with the corresponding Genus name
Tax[is.na(Tax$Species),"Species"] <- Tax[is.na(Tax$Species),"Genus"]
saveRDS(Tax, "dati/tassonomia.rds")
```

```{r}
# Join and Reclassification
tax_prep <- Tax %>% 
  rename(species = ID_Specie, Species_name = Species)

veg_prep <- left_join(df_meta, tax_prep, by = "species")
Genus_target <- c("Pinus","Quercus","Fagus")
veg <- veg_prep %>% 
  mutate(
    gruppo_climatico = case_when(
      grepl("^BS", climate_zona) ~ "B",
      grepl("^C", climate_zona) ~ "C",
      grepl("^D",  climate_zona) ~ "D",
      grepl("^ET", climate_zona) ~ "E")) %>%
  filter(Genus %in% Genus_target) %>%
  rename(gruppo_specie = Genus)

# Final Saving of meta_essential
df_veg <- veg[, c("stand_id", "gruppo_climatico", "gruppo_specie", "longitude", "latitude")]
saveRDS(df_veg, "dati/meta_group.rds")
```

## pixel extraction

 - File esterni
meta_group.rds
CRU_TS_pre.nc

```{r}
library(ncdf4)
```

```{r}
df_veg <- readRDS("dati/meta_group.rds")
nc_data <- nc_open("dati/CRU_TS_pre.nc")
map_lon <- ncvar_get(nc_data, "lon")
map_lat <- ncvar_get(nc_data, "lat")
nc_close(nc_data)
```

```{r}
find_pixel_index <- function(lon, lat, map_lon, map_lat) {
  idx_lon <- which.min(abs(map_lon - lon))
  idx_lat <- which.min(abs(map_lat - lat))
  return(paste(idx_lon, idx_lat, sep = "_"))}
```

```{r}
df_veg$pixel_id <- mapply(find_pixel_index, 
                          df_veg$longitude, 
                          df_veg$latitude, 
                          MoreArgs = list(map_lon = map_lon, map_lat = map_lat))
saveRDS(df_veg, "dati/meta_essential.rds")
```

## Clima creation

 - Oggetti esterni
df_tax — Dataframe di partenza che deve contenere almeno le colonne: pixel_id, longitude e latitude
 - File esterni
spei01.nc
CRU_TS_tmp.nc
CRU_TS_tmn.nc
CRU_TS_pre.nc
CRU_TS_pet.nc

```{r}
library(dplyr)
library(terra)
```

```{r}
df_veg <- readRDS("dati/meta_essential.rds")
df_pixel <- df_veg %>%
  distinct(pixel_id, .keep_all = TRUE) %>%
  select(pixel_id, longitude, latitude)
clima <- estrazione_clima(df_pixel)
saveRDS(clima, "dati/lista_clima.Rds")
```

# Cross Correlation

```{r}
correlazione_significativa_lag(df_veg,sel_orders = "Quercus")
correlazione_significativa_lag(df_veg,sel_orders = "Pinus")
correlazione_significativa_lag(df_veg,sel_orders = "Fagus")
```


# DLNM Preparation

```{r}
library(dplyr)
library(stringr)
library(lubridate)
```

```{r}
df_veg <- readRDS("dati/meta_essential.rds")
```


```{r}
dlnm_spei_fagus <- prepara_dlnm_misto(df_veg, var_meteo = "spei", var_ctrl = "tmn",
                                     month_n = 9, g_clima = NULL,
                                     g_specie = "Fagus", lag_n = 21)
dlnm_spei_quercus <- prepara_dlnm_misto(df_veg, var_meteo = "spei", var_ctrl = "tmn",
                                     month_n = 9, g_clima = NULL,
                                     g_specie = "Fagus", lag_n = 21)
dlnm_spei_pinus <- prepara_dlnm_misto(df_veg, var_meteo = "spei", var_ctrl = "tmn",
                                     month_n = 9, g_clima = NULL,
                                     g_specie = "Fagus", lag_n = 21)
```

```{r}
saveRDS(dlnm_spei_fagus,"dati/dlnm_21spei_all_tmn_Fagus.rds")
saveRDS(dlnm_spei_quercus,"dati/dlnm_21spei_all_tmn_Fagus.rds")
saveRDS(dlnm_spei_pinus,"dati/dlnm_21spei_all_tmn_Fagus.rds")
```

```{r}
faggio <- readRDS("dati/dlnm_21spei_all_tmn_Fagus.rds")
quercia <- readRDS("dati/dlnm_21spei_all_tmn_Quercus.rds")
pino <- readRDS("dati/dlnm_21spei_all_tmn_Pinus.rds")
```

```{r}
faggio_auto <- termine_auto(faggio$df, faggio$mat)
controllo(faggio_auto, report = FALSE, rwl_prev = TRUE)
faggio_auto$df$ID <- as.factor(faggio_auto$df$ID)
quercia_auto <- termine_auto(quercia$df, quercia$mat)
controllo(quercia_auto, report = FALSE, rwl_prev = TRUE)
quercia_auto$df$ID <- as.factor(quercia_auto$df$ID)
pino_auto <- termine_auto(pino$df, pino$mat)
controllo(pino_auto, report = FALSE, rwl_prev = TRUE)
pino_auto$df$ID <- as.factor(pino_auto$df$ID)
```

# Running model

```{r}
library(dlnm)
library(mgcv)
library(splines)
```

```{r}
cb_faggio <- crossbasis(faggio_auto$mat, lag = 20,
                       argvar = list(fun = "ns", df = 2), 
                       arglag = list(fun = "ns", df = 6))
model_faggio <- bam(rwl ~ cb_faggio + rwl_prev +
                   s(ID, bs = "re") +
                   s(anno, bs = "ps", k = 20),
                   family = Gamma(link = "log"), 
                   data = faggio_auto$df,
                   method = "fREML",
                   discrete = TRUE)
pred_faggio <- get_dlnm_pred(model_faggio, cb_faggio, cb_name = "cb_faggio")
```

```{r}
cb_quercia <- crossbasis(quercia_auto$mat, lag = 20,
                       argvar = list(fun = "ns", df = 2), 
                       arglag = list(fun = "ns", df = 6))
model_quercia <- bam(rwl ~ cb_quercia + rwl_prev +
                   s(ID, bs = "re") +
                   s(anno, bs = "ps", k = 20),
                   family = Gamma(link = "log"), 
                   data = quercia_auto$df,
                   method = "fREML",
                   discrete = TRUE)
pred_quercia <- get_dlnm_pred(model_quercia, cb_quercia, cb_name = "cb_quercia")
```

```{r}
cb_pino <- crossbasis(pino_auto$mat, lag = 20,
                       argvar = list(fun = "ns", df = 2), 
                       arglag = list(fun = "ns", df = 6))
model_pino <- bam(rwl ~ cb_pino + rwl_prev +
                   s(ID, bs = "re") +
                   s(anno, bs = "ps", k = 20),
                   family = Gamma(link = "log"), 
                   data = pino_auto$df,
                   method = "fREML",
                   discrete = TRUE)
pred_pino <- get_dlnm_pred(model_pino, cb_pino, cb_name = "cb_pino")
```

# Showing charts

```{r}
library(ggplot2)
```

```{r}
mesi_sin <- c("sep", "aug", "jul", "jun", "may", "apr", "mar", "feb", "jan") 
mesi_cos <- c("dec", "nov", "oct")
mesi_eng <- c()
for (anno in c(0:2))
{mesi_eng <- c(mesi_eng,str_replace_all(paste(mesi_sin,"-",anno), " ", ""),
                        str_replace_all(paste(mesi_cos,"-",anno+1), " ", ""))}
```

```{r}
nomi <- c("Pinus","Quercus","Fagus")
confronto_lag(lista_pred = list(pred_pino,pred_quercia,pred_faggio),
              valore_spei = 2.5, etichette_mesi = mesi_eng, lag_n=21,
              nomi_gruppi = nomi)
ggsave(filename = "gasparrini_genus_pos.pdf",
       device = "pdf",width = 7, height = 5,  units = "in",dpi = 300 )
```

```{r}
png("gasparrini_genus_neg.png", width = 7, height = 5, units = "in", res = 300)
print(confronto_lag(lista_pred = list(pred_pino,pred_quercia,pred_faggio),
              valore_spei = -2.5, etichette_mesi = mesi_eng, lag_n=21,
              nomi_gruppi = nomi))
dev.off()
```

```{r}
grafico_unito(pred_pino,pred_quercia,pred_faggio, mesi=mesi_eng, nomi=nomi,
              titolo="Lagged Effects of SPEI on Tree Growth by Genus")
ggsave(filename = "gasparrini_genus.pdf", 
       device = "pdf",width = 7, height = 10,  units = "in", dpi = 300 )
```


```{r}
png("gasparrini_genus.png", width = 7, height = 6, units = "in", res = 300)
print(grafico_unito(pred_pino,pred_quercia,pred_faggio, mesi=mesi_eng, nomi=nomi,
              titolo="Lagged Effects of SPEI on Tree Growth by Genus"))
dev.off()
```

